---
title: "Activity 2 — Task 4: Continuous Regression Modeling"
author: "Group 08"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This section addresses the second research question for Activity 2: *How accurately can we predict an employee's rating of their employer's mental health support based on age, gender, coworker reaction, and physical health support? Among these, which factor has the strongest influence?*

Because the response variable (`Employer_Mental_Importance`) operates on a strictly bounded 0–10 scale containing exact zeros, a log or Box-Cox transformation is mathematically inappropriate without arbitrary zero-shifting. Therefore, we preserve the original, highly interpretable 0–10 scale, utilizing Heteroscedasticity-Consistent (HC) robust standard errors to correct for any violations of variance assumptions.

Four model specifications are compared on a held-out test set: a **full multiple linear regression**, a parsimonious model reduced by **stepwise AIC selection**, and regularized **Ridge** and **LASSO** regressions to rigorously rank feature importance.

```{r}
library(tidyverse)
library(glmnet)
library(car)
library(lmtest)
library(sandwich)  # HC-robust standard errors
library(broom)     # tidy coefficient tables

data <- read.csv("dataset/osmi_clean_combined.csv")
set.seed(2026)     # fixed seed for reproducible data splitting
```

# 4 Continuous Regression Modeling

## 4.1 Training and Testing Partitions

Model quality is judged on observations not seen during fitting. We perform a single global 80:20 random split of the 1,055 cleaned records, assigning 844 rows to the training set and 211 to the test set.

```{r}
# Isolate the explicit variables requested by the research question
used_cols <- c("Employer_Mental_Importance", "Age", "Gender_Category", 
               "Coworker_Reaction_Score", "Employer_Phys_Importance",
               "Age_scaled", "Coworker_Reaction_scaled", "Employer_Phys_scaled")

data$Gender_Category <- relevel(factor(data$Gender_Category), ref = "Male")

# 80:20 Random Split
train_idx <- sample(nrow(data), floor(0.8 * nrow(data)))
train <- data[train_idx, ]
test  <- data[-train_idx, ]

stopifnot(nrow(train) + nrow(test) == nrow(data))

cat("Training set:", nrow(train), "observations\n")
cat("Test set:    ", nrow(test), "observations\n")
```

## 4.2 Multiple Linear Regression

We begin with an ordinary linear model regressing the mental health rating on the four core predictors.

### A reference model on the raw 0-10 scale

```{r}
ols_full <- lm(Employer_Mental_Importance ~ Age + Gender_Category + 
                 Coworker_Reaction_Score + Employer_Phys_Importance, 
               data = train)

summary(ols_full)
```

The baseline model explains approximately `r round(summary(ols_full)$adj.r.squared * 100, 1)`% of the variance in mental health ratings ($R^2_{\text{adj}} \approx$ `r round(summary(ols_full)$adj.r.squared, 3)`) and is highly significant overall ($p < 10^{-16}$). Before interpreting the coefficients, we must evaluate the underlying assumptions.

### Diagnostic checking

```{r}
# 1. Shapiro-Wilk test for Normality
sw <- shapiro.test(resid(ols_full))

# 2. Breusch-Pagan test for Homoscedasticity
bp <- lmtest::bptest(ols_full)

diag_table <- tibble(
  Test = c("Shapiro-Wilk (Normality)", "Breusch-Pagan (Homoscedasticity)"),
  Statistic = c(round(sw$statistic, 3), round(unname(bp$statistic), 1)),
  `P-value` = c(signif(sw$p.value, 3), signif(bp$p.value, 3))
)

knitr::kable(diag_table, caption = "OLS Assumption Diagnostics")
```

Both classical assumptions fail. Bounded rating scales (0-10) inherently generate non-normal residuals with dense clustering. Furthermore, the strong failure of the Breusch-Pagan test indicates heteroscedasticity (unequal error variances). To ensure our conclusions are not artifacts of these violations, we will apply HC-robust standard errors.

### Refining the model by AIC (Stepwise Selection)

Are demographic variables truly necessary for predicting corporate mental health priorities? Stepwise selection by Akaike Information Criterion (AIC) objectively drops variables that do not contribute meaningful predictive power.

```{r}
ols_step <- step(ols_full, direction = "both", trace = FALSE)
summary(ols_step)

dropped_terms <- setdiff(attr(terms(ols_full), "term.labels"),
                         attr(terms(ols_step), "term.labels"))

cat("\nTerms dropped by AIC:", paste(dropped_terms, collapse = ", "), "\n")
```

The AIC algorithm entirely eliminates **Age** and **Gender_Category**. The parsimonious model predicts equally well using only the two workplace environmental factors, proving that personal demographics contain almost no predictive signal for this specific rating.

### Robust Inference and Effect Sizes

Because the model failed the Breusch-Pagan test, we re-evaluate the stepwise-selected coefficients using HC1-robust standard errors.

```{r}
robust_se <- lmtest::coeftest(ols_step, vcov = sandwich::vcovHC(ols_step, type = "HC1"))
print(robust_se)
```

Even after correcting for unequal variances, both workplace factors remain highly significant ($p < 0.001$).

**What drives the mental health rating?**

Based on the raw coefficients:

1.  **Physical Health Priority (Strongest Driver):** For every 1-point increase in an employer's physical health rating, the mental health rating rises by roughly +`r round(coef(ols_step)["Employer_Phys_Importance"], 2)` points. Companies that care about physical well-being overwhelmingly tend to care about mental well-being.

2.  **Coworker Reaction (Secondary Driver):** For every 1-point increase in perceived coworker openness, the mental health rating rises by roughly +`r round(coef(ols_step)["Coworker_Reaction_Score"], 2)` points. A supportive peer environment moderately signals a supportive corporate environment.

## 4.3 Ridge and LASSO Regression (Standardized Importance)

While raw coefficients are interpretable, comparing a 1-unit jump in two different variables can be slightly misleading. Shrinkage models (Ridge and LASSO) require standardized inputs (Z-scores), forcing every predictor onto a mathematically level playing field. The resulting coefficients objectively rank "which factor has the strongest influence."

```{r}
# Build matrices using the pre-scaled Z-score variables from Task 1
x_train <- model.matrix(~ Age_scaled + Gender_Category + 
                          Coworker_Reaction_scaled + Employer_Phys_scaled, data = train)[, -1]
y_train <- train$Employer_Mental_Importance

x_test <- model.matrix(~ Age_scaled + Gender_Category + 
                         Coworker_Reaction_scaled + Employer_Phys_scaled, data = test)[, -1]
y_test <- test$Employer_Mental_Importance

# Cross-validated Ridge and LASSO
set.seed(2026)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

lasso_coefs <- as.matrix(coef(cv_lasso, s = "lambda.min"))
cat("\nLASSO Standardized Coefficients (lambda.min):\n")
print(round(lasso_coefs[lasso_coefs[,1] != 0, , drop=FALSE], 3))
```

LASSO entirely zeroes out `Age_scaled` and the gender dummy `Female`, exactly mirroring the Stepwise AIC logic.

Looking at the standardized coefficients, Employer_Phys_scaled ($\approx$ `r round(lasso_coefs["Employer_Phys_scaled", 1], 2)`) is nearly twice as large as Coworker_Reaction_scaled ($\approx$ `r round(lasso_coefs["Coworker_Reaction_scaled", 1], 2)`). This definitively answers the research question: Employer focus on physical health is by far the strongest indicator of their focus on mental health, exerting nearly twice the predictive influence of peer culture.

## 4.4 Predictive Performance on the Held-out Data

We score all four models on the 211 untouched test records to evaluate their out-of-sample accuracy (Root Mean Squared Error and $R^2$).

```{r}
pred_full  <- predict(ols_full, newdata = test)
pred_step  <- predict(ols_step, newdata = test)
pred_ridge <- as.numeric(predict(cv_ridge, newx = x_test, s = "lambda.min"))
pred_lasso <- as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.min"))

score_model <- function(name, preds, actual) {
  err <- preds - actual
  rmse <- sqrt(mean(err^2))
  mae <- mean(abs(err))
  r2 <- 1 - sum(err^2) / sum((actual - mean(actual))^2)
  tibble(Model = name, RMSE = rmse, MAE = mae, `Test R-squared` = r2)
}

comparison <- bind_rows(
  score_model("Full OLS", pred_full, y_test),
  score_model("Stepwise OLS", pred_step, y_test),
  score_model("Ridge", pred_ridge, y_test),
  score_model("LASSO", pred_lasso, y_test)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(comparison %>% arrange(RMSE), caption = "Out-of-Sample Performance (0-10 Scale)")
```

### Conclusion for Research Question 2

**1. Predictive Accuracy:** We can predict an employee's rating of their employer's mental health support with an average error margin (RMSE) of roughly `r round(comparison$RMSE[comparison$Model == "Stepwise OLS"], 2)` points on the 0-10 scale. The models successfully explain about `r round(comparison[["Test R-squared"]][comparison$Model == "Stepwise OLS"] * 100, 1)`% of the variance in completely unseen data. The parsimonious Stepwise model and the LASSO model performed slightly better than the full model, proving that complex demographics are not needed.

**2. Strongest Driving Factor:** Both the raw stepwise coefficients and the standardized LASSO penalty confirm that **an employer's prioritization of physical health** is the overwhelmingly dominant predictor. It exerts twice the influence of coworker culture, while personal demographics (Age and Gender) hold zero statistical power in predicting how an employer's support network is rated.
