---
title: 'Activity 2 — Task 5: Binary Logistic Regression Modeling'
author: "Group 08"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 5 Binary Logistic Regression Modeling

## 5.1 Research Question and Data Preparation

This final task addresses **Research Question 2 (RQ2)** for Activity 2:
**Which personal factors (age, gender, family history) and environmental
factors (employer prioritization, coworker reactions, and technology-company
status) affect the probability of seeking professional mental-health treatment,
and how accurately can a model classify treatment-seeking respondents?**

Because the target variable (`Treatment_Cat`) is binary (Yes/No), we employ a Binary Logistic Regression model.

To maintain strict comparability with the continuous regression analysis, we utilize the exact same globally randomized 80:20 split (seed 2026).

```{r}
library(tidyverse)
library(caret)
library(pROC)
library(broom)

data <- read.csv("dataset/osmi_clean_combined.csv")

# Ensure reference levels are set logically for interpretation
data <- data %>%
  mutate(
    Treatment_Cat = factor(Treatment_Cat, levels = c("No", "Yes")),
    Gender_Category = relevel(factor(Gender_Category), ref = "Male"),
    Family_History = relevel(factor(Family_History), ref = "I don't know"),
    Tech_Company_Cat = relevel(factor(Tech_Company_Cat), ref = "No")
  )

set.seed(2026)
train_idx <- sample(nrow(data), floor(0.8 * nrow(data)))
train <- data[train_idx, ]
test  <- data[-train_idx, ]

cat("Training cases:", nrow(train), "| Test cases:", nrow(test), "\n")
```

## 5.2 Model Fitting and Odds Ratios

We fit the logistic regression model on the training set using the seven
specified demographic and workplace predictor groups. The reference groups are
No treatment, Male gender, unknown family history, and a non-technology
company.

```{r}
logit_model <- glm(
  Treatment_Cat ~ Age + Gender_Category + Family_History + 
    Employer_Mental_Importance + Employer_Phys_Importance + 
    Coworker_Reaction_Score + Tech_Company_Cat,
  family = binomial(link = "logit"),
  data = train
)

# Extract coefficients, odds ratios, p-values, and 95% Wald confidence
# intervals directly from the fitted model.
wald_ci <- confint.default(logit_model)
model_tidy <- broom::tidy(logit_model)

odds_ratios <- model_tidy %>%
  filter(term != "(Intercept)") %>%
  mutate(
    conf_low = wald_ci[term, 1],
    conf_high = wald_ci[term, 2],
    Predictor = dplyr::case_when(
      term == "Age" ~ "Age",
      term == "Gender_CategoryFemale" ~ "Gender: Female",
      term == "Gender_CategoryOther" ~ "Gender: Other",
      term == "Family_HistoryNo" ~ "Family history: No",
      term == "Family_HistoryYes" ~ "Family history: Yes",
      term == "Employer_Mental_Importance" ~ "Employer mental importance",
      term == "Employer_Phys_Importance" ~ "Employer physical importance",
      term == "Coworker_Reaction_Score" ~ "Coworker reaction score",
      term == "Tech_Company_CatYes" ~ "Technology company",
      TRUE ~ term
    ),
    `Odds Ratio (OR)` = exp(estimate),
    `95% CI for OR` = sprintf("%.3f to %.3f", exp(conf_low), exp(conf_high)),
    Significant = p.value < 0.05,
    Estimate = round(estimate, 3),
    `Odds Ratio (OR)` = round(`Odds Ratio (OR)`, 3),
    `P-value` = if_else(p.value < 0.001, "< 0.001", sprintf("%.3f", p.value))
  ) %>%
  select(Predictor, Estimate, `Odds Ratio (OR)`, `95% CI for OR`, `P-value`, Significant)

knitr::kable(odds_ratios, caption = "Adjusted Odds Ratios for Seeking Treatment")

term_stat <- function(model_term, statistic) {
  model_tidy %>% filter(.data$term == model_term) %>% pull({{ statistic }})
}
fmt3 <- function(x) format(round(x, 3), nsmall = 3)

age_or <- exp(term_stat("Age", estimate))
age_p <- term_stat("Age", p.value)
female_or <- exp(term_stat("Gender_CategoryFemale", estimate))
female_p <- term_stat("Gender_CategoryFemale", p.value)
other_or <- exp(term_stat("Gender_CategoryOther", estimate))
other_p <- term_stat("Gender_CategoryOther", p.value)
family_no_beta <- term_stat("Family_HistoryNo", estimate)
family_no_or <- exp(family_no_beta)
family_no_p <- term_stat("Family_HistoryNo", p.value)
family_yes_beta <- term_stat("Family_HistoryYes", estimate)
family_yes_or <- exp(family_yes_beta)
family_yes_p <- term_stat("Family_HistoryYes", p.value)
family_yes_vs_no_or <- family_yes_or / family_no_or
```

### What drives the decision to seek treatment?

By examining the **Odds Ratios (OR)**, where $OR > 1$ indicates increased probability and $OR < 1$ indicates decreased probability, we uncover a fascinating divide between personal and environmental factors:

- **Family History Is the Strongest Predictor:** Relative to respondents who
  answered "I don't know," respondents answering No have substantially lower
  odds of seeking treatment ($\beta \approx$ `r fmt3(family_no_beta)`,
  $OR \approx$ `r fmt3(family_no_or)`, $p$ `r if (family_no_p < 0.001) "< 0.001" else paste0("= ", fmt3(family_no_p))`), while those answering Yes have substantially higher odds
  ($\beta \approx$ `r fmt3(family_yes_beta)`, $OR \approx$ `r fmt3(family_yes_or)`,
  $p$ `r if (family_yes_p < 0.001) "< 0.001" else paste0("= ", fmt3(family_yes_p))`). The implied
  adjusted Yes-versus-No odds ratio is approximately `r fmt3(family_yes_or)` /
  `r fmt3(family_no_or)` = `r fmt3(family_yes_vs_no_or)`.
  Age also has a small, statistically significant positive association
  ($OR \approx$ `r fmt3(age_or)`, $p = `r fmt3(age_p)`): each additional year is associated with
  about `r round((age_or - 1) * 100, 1)`% increase in the odds of seeking treatment, holding the other
  variables constant. Female respondents have significantly higher odds than
  Male respondents ($OR \approx$ `r fmt3(female_or)`, $p = `r fmt3(female_p)`);
  the Other category does not significantly differ from Male ($OR \approx
  `r fmt3(other_or)`, $p = `r fmt3(other_p)`).

- **Workplace Factors Are Not Statistically Significant:** Employer
  mental-health importance, employer physical-health importance, coworker
  reaction score, and technology-company status are not statistically
  significant at the 5% level in this model.

In short: **family history is the dominant predictor, with smaller positive
associations for age and Female gender; the measured workplace/environmental
factors do not show statistically significant associations at the 5% level.**

## 5.3 Out-of-Sample Classification Performance

To answer the second half of the research question ("How accurately can the model classify?"), we apply the fitted model to the 211 unseen records in the test set.

```{r}
# Generate probabilities and classes on the Test Set
test_probs <- predict(logit_model, newdata = test, type = "response")
test_preds <- factor(ifelse(test_probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
actual_class <- test$Treatment_Cat

# Confusion Matrix
conf_matrix <- confusionMatrix(test_preds, actual_class, positive = "Yes")
print(conf_matrix$table)

# Extract Metrics for inline text
acc_val <- conf_matrix$overall["Accuracy"]
prec_val <- conf_matrix$byClass["Pos Pred Value"]
rec_val <- conf_matrix$byClass["Sensitivity"]

# ROC and AUC
roc_curve <- roc(actual_class, test_probs)
auc_value <- as.numeric(auc(roc_curve))

cat("\nAccuracy: ", round(acc_val, 3))
cat("\nPrecision:", round(prec_val, 3))
cat("\nRecall:   ", round(rec_val, 3))
cat("\nROC-AUC:  ", round(auc_value, 3), "\n")

plot(roc_curve, col = "firebrick", lwd = 2, 
     main = paste("ROC Curve (Test Set) - AUC =", round(auc_value, 3)))
```

### Conclusion for Research Question 2

The model classifies treatment-seeking respondents with an overall **Accuracy
of `r round(acc_val * 100, 1)`%** on the 211 completely unseen test
observations.

More importantly, the **ROC-AUC score is `r round(auc_value, 3)`**. This means that if we randomly select one employee who sought treatment and one who did not, the model has an `r round(auc_value * 100, 1)`% probability of correctly ranking the treatment-seeker with a higher risk score.

The precision (`r round(prec_val, 3)`) and recall (`r round(rec_val, 3)`) show
that the classifier identifies treatment-seeking respondents reasonably well at
the 0.5 probability threshold.

**Final Answer:** The logistic model achieves approximately `r round(acc_val * 100, 1)`% accuracy and a ROC-AUC of `r round(auc_value, 3)`, indicating moderate discrimination. Family history is the strongest predictor; age has a small but statistically significant positive association, and Female respondents have significantly higher odds of seeking treatment than Male respondents. The workplace/environmental variables in this model are not statistically significant at the 5% level.

## 5.4 Random Forest Benchmark

The logistic model is linear and assumes log-odds are a linear function of the predictors. To check whether a *nonlinear* tree ensemble can capture additional structure---interactions between predictors or curvilinear effects---we fit a **Random Forest** classifier on the exact same binary target (`Treatment_Cat`) and the same training/test partition (seed 2026, 844/211). Random Forest aggregates many decision trees, each trained on a bootstrap sample with a random subset of predictors, which lets it model nonlinearity and interactions without parametric assumptions.

```{r}
# Random Forest needs its own package
library(randomForest)

# Numeric design matrix (same predictors as the logistic model) for x/y interface
rf_X <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = train
)[, -1]
rf_y <- train$Treatment_Cat

# ---- 5-fold stratified cross-validation to tune the random forest ----
set.seed(2026)
folds <- createFolds(rf_y, k = 5, list = TRUE, returnTrain = TRUE)

grid <- expand.grid(
  mtry     = c(2, 3, 5),
  maxnodes = c(8, 16, 32),
  ntree    = c(100, 300)
)

cv_results <- data.frame()
for (g in seq_len(nrow(grid))) {
  aucs <- numeric(5)
  for (f in seq_along(folds)) {
    tr_idx <- folds[[f]]
    set.seed(123)
    rf_f <- randomForest(
      x = rf_X[tr_idx, , drop = FALSE], y = rf_y[tr_idx],
      mtry = grid$mtry[g], maxnodes = grid$maxnodes[g], ntree = grid$ntree[g]
    )
    p_f <- predict(rf_f, rf_X[-tr_idx, , drop = FALSE], type = "prob")[, "Yes"]
    aucs[f] <- as.numeric(auc(roc(rf_y[-tr_idx], p_f)))
  }
  cv_results <- rbind(cv_results, data.frame(
    mtry = grid$mtry[g], maxnodes = grid$maxnodes[g],
    ntree = grid$ntree[g], CV_AUC = round(mean(aucs), 4)
  ))
}
best_cfg <- cv_results[which.max(cv_results$CV_AUC), ]
knitr::kable(cv_results[order(-cv_results$CV_AUC), ],
             caption = "5-fold CV tuning of the random forest (mean validation AUC)")

# ---- Retrain best config on the full training set and score the test set ----
set.seed(123)
rf_model <- randomForest(
  x = rf_X, y = rf_y,
  mtry = best_cfg$mtry, maxnodes = best_cfg$maxnodes,
  ntree = best_cfg$ntree, importance = TRUE
)

# Out-of-bag error rate
oob_err <- mean(rf_model$err.rate[, 1])

# Predict on the held-out test set (use the same model.matrix structure)
rf_X_test <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = test
)[, -1]
rf_probs <- predict(rf_model, rf_X_test, type = "prob")[, "Yes"]
rf_preds <- factor(ifelse(rf_probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))

rf_cm <- confusionMatrix(rf_preds, test$Treatment_Cat, positive = "Yes")
print(rf_cm$table)

# Metrics
rf_acc  <- rf_cm$overall["Accuracy"]
rf_prec <- rf_cm$byClass["Precision"]
rf_rec  <- rf_cm$byClass["Sensitivity"]
rf_f1   <- rf_cm$byClass["F1"]
rf_auc  <- as.numeric(auc(roc(test$Treatment_Cat, rf_probs)))

cat("\nBest config (5-fold CV): mtry =", best_cfg$mtry,
    ", maxnodes =", best_cfg$maxnodes, ", ntree =", best_cfg$ntree)
cat("\nRandom Forest OOB error: ", round(oob_err, 3))
cat("\nAccuracy: ", round(rf_acc, 3))
cat("\nPrecision:", round(rf_prec, 3))
cat("\nRecall:   ", round(rf_rec, 3))
cat("\nF1-Score: ", round(rf_f1, 3))
cat("\nROC-AUC:  ", round(rf_auc, 3), "\n")
```

```{r}
# Variable importance: how much each predictor reduces classification accuracy.
# Aggregate dummy-level rows back to their original predictors.
raw_imp <- importance(rf_model)[, "MeanDecreaseAccuracy", drop = FALSE]
base_names <- sub('(Gender_Category|Family_History|Tech_Company_Cat).*', '\\1',
                  rownames(raw_imp))
imp_by_predictor <- rowsum(raw_imp[, 1], base_names, reorder = FALSE)
rf_imp <- imp_by_predictor[, 1]
names(rf_imp) <- rownames(imp_by_predictor)

rf_imp_df <- data.frame(
  Predictor = names(rf_imp),
  MeanDecreaseAccuracy = round(rf_imp, 2)
) %>%
  arrange(desc(MeanDecreaseAccuracy))
knitr::kable(rf_imp_df, caption = "Random Forest Variable Importance (Mean Decrease in Accuracy)")
```

```{r}
# Comparison of the two classifiers on the held-out test set
logit_f1 <- 2 * prec_val * rec_val / (prec_val + rec_val)
comparison_rf <- tibble(
  Model = c("Binary Logistic Regression", "Random Forest"),
  Accuracy  = c(round(acc_val, 3), round(rf_acc, 3)),
  Precision = c(round(prec_val, 3), round(rf_prec, 3)),
  Recall    = c(round(rec_val, 3), round(rf_rec, 3)),
  `F1-Score` = c(round(logit_f1, 3), round(rf_f1, 3)),
  `ROC-AUC` = c(round(auc_value, 3), round(rf_auc, 3))
)
knitr::kable(comparison_rf, caption = "Test-Set Comparison: Logistic vs. Random Forest")
```

### Random Forest Benchmark Results

The Random Forest (hyperparameters chosen by 5-fold cross-validation: `mtry = `r best_cfg$mtry`, `maxnodes = `r best_cfg$maxnodes`, `ntree = `r best_cfg$ntree`) achieves an out-of-sample **accuracy of `r round(rf_acc * 100, 1)`% with a ROC-AUC of `r round(rf_auc, 3)`** (`r round(rf_f1, 3)` F1), compared with the logistic model's accuracy of `r round(acc_val * 100, 1)`% and AUC of `r round(auc_value, 3)`. The two classifiers perform very similarly, with the linear logistic model holding a slight edge on this small tabular sample---consistent with the expectation that a flexible tree ensemble has little extra signal to extract from only 844 training rows.

Critically, the Random Forest's variable-importance ranking **corroborates the logistic finding**:            `Family_History` is far and away the most influential predictor (mean decrease in accuracy $\approx$ `r round(rf_imp["Family_History"], 0)`), while the workplace measures, tech-company status, and gender contribute only marginal splits. Whether modeled linearly or nonlinearly, family history dominates the decision to seek treatment. The one nuance is that the linear logistic model detects a modest female-versus-male odds advantage ($OR=1.644$) that the partitions of this small tree ensemble do not exploit.

## 5.5 Principal Component Analysis (PCA) with Random Forest

The predictors include several correlated workplace ratings and demographic measures spanning different scales (Age 19--65, ratings 0--10, binary indicators). **Principal Component Analysis (PCA)** decorrelates these features and compresses them into a smaller set of orthogonal components ordered by the variance they explain. This is a common preprocessing step for reducing multicollinearity and dimensionality. For a Random Forest, we test empirically whether feeding the model PCA components---rather than the original correlated features---changes predictive performance.

The transformation is fitted **only on the training set** (standardizing each feature and extracting the principal-component loadings) and then applied to the test set, avoiding data leakage.

```{r}
# Build the numeric design matrix (same predictors as the logistic model);
# one-hot encode the categorical predictors, dropping base levels as in model.matrix
X_all <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = rbind(train, test)
)[, -1]

n_train <- nrow(train)
X_train_mat <- X_all[1:n_train, ]
X_test_mat  <- X_all[(n_train + 1):nrow(X_all), ]

# PCA on the TRAINING set only (center + scale), then transform test
pca_fit <- prcomp(X_train_mat, center = TRUE, scale. = TRUE)

# Proportion of variance explained by the first k components
cum_var <- cumsum(pca_fit$sdev^2 / sum(pca_fit$sdev^2))

# Transform both sets using the training PCA loadings
Z_train <- predict(pca_fit, newdata = X_train_mat)
Z_test  <- predict(pca_fit, newdata = X_test_mat)

# Variance explained
pca_summary <- tibble(
  Component = 1:length(cum_var),
  `Prop. Variance` = round(pca_fit$sdev^2 / sum(pca_fit$sdev^2), 3),
  `Cumulative`      = round(cum_var, 3)
)
knitr::kable(pca_summary, caption = "Variance explained by PCA components")
```

```{r}
# ---- 5-fold stratified CV to tune the PCA random forest (k = 8 components) ----
k_pca <- 8   # drop the lowest-variance component from the 9-feature design matrix

set.seed(2026)
pca_folds <- createFolds(train$Treatment_Cat, k = 5, list = TRUE, returnTrain = TRUE)

pca_grid <- expand.grid(
  mtry     = c(1, 2, 3),
  maxnodes = c(8, 16, 32),
  ntree    = c(100, 300)
)

pca_cv_results <- data.frame()
for (g in seq_len(nrow(pca_grid))) {
  aucs <- numeric(5)
  for (f in seq_along(pca_folds)) {
    tr_idx <- pca_folds[[f]]
    set.seed(123)
    rf_pca_f <- randomForest(
      x = Z_train[tr_idx, 1:k_pca, drop = FALSE], y = train$Treatment_Cat[tr_idx],
      mtry = pca_grid$mtry[g], maxnodes = pca_grid$maxnodes[g],
      ntree = pca_grid$ntree[g]
    )
    p_f <- predict(rf_pca_f, Z_train[-tr_idx, 1:k_pca, drop = FALSE], type = "prob")[, "Yes"]
    aucs[f] <- as.numeric(auc(roc(train$Treatment_Cat[-tr_idx], p_f)))
  }
  pca_cv_results <- rbind(pca_cv_results, data.frame(
    mtry = pca_grid$mtry[g], maxnodes = pca_grid$maxnodes[g],
    ntree = pca_grid$ntree[g], CV_AUC = round(mean(aucs), 4)
  ))
}
pca_best_cfg <- pca_cv_results[which.max(pca_cv_results$CV_AUC), ]
knitr::kable(pca_cv_results[order(-pca_cv_results$CV_AUC), ],
             caption = "5-fold CV tuning of the PCA random forest (k = 8 components, mean validation AUC)")

# ---- Retrain best PCA config on the full training set, score the test set ----
set.seed(123)
rf_pca <- randomForest(
  x = Z_train[, 1:k_pca, drop = FALSE], y = train$Treatment_Cat,
  mtry = pca_best_cfg$mtry, maxnodes = pca_best_cfg$maxnodes,
  ntree = pca_best_cfg$ntree
)

p_pca_rf <- predict(rf_pca, Z_test[, 1:k_pca, drop = FALSE], type = "prob")[, "Yes"]
cl_pca_rf <- factor(ifelse(p_pca_rf >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
cm_pca_rf <- confusionMatrix(cl_pca_rf, test$Treatment_Cat, positive = "Yes")

pca_rf_auc  <- as.numeric(auc(roc(test$Treatment_Cat, p_pca_rf)))
pca_rf_acc  <- cm_pca_rf$overall["Accuracy"]
pca_rf_prec <- cm_pca_rf$byClass["Precision"]
pca_rf_rec  <- cm_pca_rf$byClass["Sensitivity"]
pca_rf_f1   <- cm_pca_rf$byClass["F1"]

cat("\nBest PCA config (5-fold CV): mtry =", pca_best_cfg$mtry,
    ", maxnodes =", pca_best_cfg$maxnodes, ", ntree =", pca_best_cfg$ntree)
cat("\nPCA-RF (k=8) Accuracy:", round(pca_rf_acc, 3),
    " Precision:", round(pca_rf_prec, 3),
    " Recall:", round(pca_rf_rec, 3),
    " F1:", round(pca_rf_f1, 3),
    " ROC-AUC:", round(pca_rf_auc, 3), "\n")
```

```{r}
# 3-way comparison: logistic, raw random forest, and PCA random forest
comparison_pca <- tibble(
  Model = c("Binary Logistic", "Random Forest (raw)", "Random Forest (PCA k=8)"),
  Accuracy  = c(round(acc_val, 3), round(rf_acc, 3), round(pca_rf_acc, 3)),
  Precision = c(round(prec_val, 3), round(rf_prec, 3), round(pca_rf_prec, 3)),
  Recall    = c(round(rec_val, 3), round(rf_rec, 3), round(pca_rf_rec, 3)),
  `F1-Score` = c(round(logit_f1, 3), round(rf_f1, 3), round(pca_rf_f1, 3)),
  `ROC-AUC` = c(round(auc_value, 3), round(rf_auc, 3), round(pca_rf_auc, 3))
)
knitr::kable(comparison_pca, caption = "Test-Set Comparison: Logistic vs. Random Forest (raw and PCA)")
```

### PCA-RF Benchmark Results

With $k=8$ PCA components (which, from the nine-predictor design matrix, retain $\approx$ `r round(cum_var[8] * 100, 1)`% of the variance by dropping the single lowest-variance component) and the same 5-fold cross-validation used for the raw forest---here selecting `mtry = `r pca_best_cfg$mtry`, `maxnodes = `r pca_best_cfg$maxnodes`, `ntree = `r pca_best_cfg$ntree`---the PCA-transformed Random Forest attains a **ROC-AUC of `r round(pca_rf_auc, 3)` and accuracy of `r round(pca_rf_acc * 100, 1)`%** on the held-out test set. On ROC-AUC it edges out both the raw-feature Random Forest and the linear logistic model (`r round(auc_value, 3)`); on accuracy it lands just below the logistic baseline. The modest gain and near-matching accuracy reflect the fact that the treatment decision is driven by one dominant binary predictor (family history) that a linear model already captures, so even a tuned nonlinear ensemble and a mild dimensionality reduction add little beyond ranking better on borderline cases. The practical takeaway is unchanged: no PCA preprocessing alters the substantive conclusion that treatment-seeking is governed by personal biology rather than the corporate environment or gender.
