---
title: 'Activity 2 — Task 5: Binary Logistic Regression Modeling'
author: "Group 08"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 5 Binary Logistic Regression Modeling

## 5.1 Research Question and Data Preparation

This final task addresses the third research question for Activity 2: **Which personal factors (age, gender, family history) and environmental factors (employer prioritization, coworker reactions) increase or decrease the probability of an employee seeking professional mental health treatment? How accurately can a model classify those who need treatment?**

Because the target variable (`Treatment_Cat`) is binary (Yes/No), we employ a Binary Logistic Regression model.

To maintain strict comparability with the continuous regression analysis, we utilize the exact same globally randomized 80:20 split (seed 2026).

```{r}
library(tidyverse)
library(caret)
library(pROC)
library(broom)

data <- read.csv("dataset/osmi_clean_combined.csv")

# Ensure reference levels are set logically for interpretation
data <- data %>%
  mutate(
    Treatment_Cat = factor(Treatment_Cat, levels = c("No", "Yes")),
    Gender_Category = relevel(factor(Gender_Category), ref = "Male"),
    Family_History = relevel(factor(Family_History), ref = "No"),
    Tech_Company_Cat = relevel(factor(Tech_Company_Cat), ref = "No")
  )

set.seed(2026)
train_idx <- sample(nrow(data), floor(0.8 * nrow(data)))
train <- data[train_idx, ]
test  <- data[-train_idx, ]

cat("Training cases:", nrow(train), "| Test cases:", nrow(test), "\n")
```

## 5.2 Model Fitting and Odds Ratios

We fit the logistic regression model on the training set using both demographic and corporate environment predictors.

```{r}
logit_model <- glm(
  Treatment_Cat ~ Age + Gender_Category + Family_History + 
    Employer_Mental_Importance + Employer_Phys_Importance + 
    Coworker_Reaction_Score + Tech_Company_Cat,
  family = binomial(link = "logit"),
  data = train
)

# Extract Odds Ratios (OR) and Confidence Intervals
model_matrix <- summary(logit_model)$coefficients
wald_ci <- confint.default(logit_model)

odds_ratios <- tibble(
  Predictor = rownames(model_matrix),
  Estimate = model_matrix[, "Estimate"],
  `Odds Ratio (OR)` = exp(Estimate),
  `P-value` = model_matrix[, "Pr(>|z|)"],
  Significant = `P-value` < 0.05
) %>%
  filter(Predictor != "(Intercept)") %>%
  mutate(
    across(c(Estimate, `Odds Ratio (OR)`), ~ round(.x, 3)),
    `P-value` = format.pval(`P-value`, digits = 3, eps = 0.001)
  )

knitr::kable(odds_ratios, caption = "Adjusted Odds Ratios for Seeking Treatment")
```

### What drives the decision to seek treatment?

By examining the **Odds Ratios (OR)**, where $OR > 1$ indicates increased probability and $OR < 1$ indicates decreased probability, we uncover a fascinating divide between personal and environmental factors:

- **Personal Family History Dominates:** The strongest predictor by a massive margin is personal biology. Employees with a known family history of mental illness have incredibly high odds of seeking treatment compared to those without. Age also shows a slight, statistically significant positive correlation ($OR = 1.024$, $p = 0.013$). Gender shows a smaller but still significant association: Female employees have higher odds of seeking treatment than Male employees ($OR = 1.644$, $p = 0.008$), while the Other category does not differ from Male.

- **Environmental Factors are Weak:** Surprisingly, workplace variables have almost zero impact on the individual's decision to seek clinical help. The importance the employer places on mental/physical health, the tech-company status, and even the coworker reaction score all yielded non-significant p-values.

In short: **The decision to seek professional treatment is driven primarily by family history, with smaller contributions from age and female gender, and essentially none from corporate policy.**

## 5.3 Out-of-Sample Classification Performance

To answer the second half of the research question ("How accurately can the model classify?"), we apply the fitted model to the 211 unseen records in the test set.

```{r}
# Generate probabilities and classes on the Test Set
test_probs <- predict(logit_model, newdata = test, type = "response")
test_preds <- factor(ifelse(test_probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
actual_class <- test$Treatment_Cat

# Confusion Matrix
conf_matrix <- confusionMatrix(test_preds, actual_class, positive = "Yes")
print(conf_matrix$table)

# Extract Metrics for inline text
acc_val <- conf_matrix$overall["Accuracy"]
prec_val <- conf_matrix$byClass["Precision"]
rec_val  <- conf_matrix$byClass["Sensitivity"]
logit_f1 <- conf_matrix$byClass["F1"]

# ROC and AUC
roc_curve <- roc(actual_class, test_probs)
auc_value <- as.numeric(auc(roc_curve))

cat("\nAccuracy: ", round(acc_val, 3))
cat("\nPrecision:", round(prec_val, 3))
cat("\nRecall:   ", round(rec_val, 3))
cat("\nF1-Score: ", round(logit_f1, 3))
cat("\nROC-AUC:  ", round(auc_value, 3), "\n")
```

```{r logit-roc-plot, fig.width=6, fig.height=5, fig.align='center'}
# Refined ROC Plot for Logistic Model
par(mar = c(4.5, 4.5, 3, 1))
plot(roc_curve, col = "firebrick", lwd = 2.5, legacy.axes = TRUE, ylim = c(0, 1),
     main = paste0("Logistic Regression ROC Curve\n(Test Set AUC = ", round(auc_value, 3), ")"),
     xlab = "False Positive Rate (1 - Specificity)", 
     ylab = "True Positive Rate (Sensitivity)", cex.main = 0.95)

abline(a = 0, b = 1, lty = 2, col = "gray50")
legend("bottomright", 
       legend = c(paste("Logistic Model (AUC =", round(auc_value, 3), ")"), "Random Chance"),
       col = c("firebrick", "gray50"), lwd = c(2.5, 1), lty = c(1, 2), bty = "n", cex = 0.85)
par(mar = c(5.1, 4.1, 4.1, 2.1))
```

### Conclusion for Research Question 3

The model successfully identifies whether an employee will seek professional mental health treatment with an overall **Accuracy of `r round(acc_val * 100, 1)`%** on completely unseen data.

More importantly, the **ROC-AUC score is `r round(auc_value, 3)`**. This means that if we randomly select one employee who sought treatment and one who did not, the model has an `r round(auc_value * 100, 1)`% probability of correctly ranking the treatment-seeker with a higher risk score.

The precision (`r round(prec_val, 3)`) and recall (`r round(rec_val, 3)`) demonstrate a well-balanced classifier that doesn't just guess "Yes" for everyone.

**Final Answer:** We can classify treatment-seekers with high accuracy ($\approx$ `r round(acc_val * 100, 1)`%). The analysis proves that the decision to seek treatment is profoundly driven by **personal biology** (specifically family history), with smaller positive associations for **age and female gender**, while corporate environmental factors (like supportive coworkers or HR policies) fail to statistically alter an employee's underlying probability of seeking professional medical care.

## 5.4 Random Forest Benchmark

The logistic model is linear and assumes log-odds are a linear function of the predictors. To check whether a *nonlinear* tree ensemble can capture additional structure---interactions between predictors or curvilinear effects---we fit a **Random Forest** classifier on the exact same binary target (`Treatment_Cat`) and the same training/test partition (seed 2026, 844/211).

```{r}
library(randomForest)

# Independent design matrix construction to prevent data leakage
rf_X <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = train
)[, -1]
rf_y <- train$Treatment_Cat

# 5-fold stratified cross-validation
set.seed(2026)
folds <- createFolds(rf_y, k = 5, list = TRUE, returnTrain = TRUE)

grid <- expand.grid(
  mtry     = c(2, 3, 5),
  maxnodes = c(8, 16, 32),
  ntree    = c(100, 300)
)

cv_results <- data.frame()
for (g in seq_len(nrow(grid))) {
  aucs <- numeric(5)
  for (f in seq_along(folds)) {
    tr_idx <- folds[[f]]
    set.seed(123)
    rf_f <- randomForest(
      x = rf_X[tr_idx, , drop = FALSE], y = rf_y[tr_idx],
      mtry = grid$mtry[g], maxnodes = grid$maxnodes[g], ntree = grid$ntree[g]
    )
    p_f <- predict(rf_f, rf_X[-tr_idx, , drop = FALSE], type = "prob")[, "Yes"]
    aucs[f] <- as.numeric(auc(roc(rf_y[-tr_idx], p_f)))
  }
  cv_results <- rbind(cv_results, data.frame(
    mtry = grid$mtry[g], maxnodes = grid$maxnodes[g],
    ntree = grid$ntree[g], CV_AUC = round(mean(aucs), 4)
  ))
}
best_cfg <- cv_results[which.max(cv_results$CV_AUC), ]
knitr::kable(cv_results[order(-cv_results$CV_AUC), ],
             caption = "5-fold CV tuning of the random forest (mean validation AUC)")

# Retrain best config on full train set
set.seed(123)
rf_model <- randomForest(
  x = rf_X, y = rf_y,
  mtry = best_cfg$mtry, maxnodes = best_cfg$maxnodes,
  ntree = best_cfg$ntree, importance = TRUE
)

oob_err <- mean(rf_model$err.rate[, 1])

# Predict on test set
rf_X_test <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = test
)[, -1]
rf_probs <- predict(rf_model, rf_X_test, type = "prob")[, "Yes"]
rf_preds <- factor(ifelse(rf_probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))

rf_cm <- confusionMatrix(rf_preds, test$Treatment_Cat, positive = "Yes")
print(rf_cm$table)

# Metrics
rf_acc  <- rf_cm$overall["Accuracy"]
rf_prec <- rf_cm$byClass["Precision"]
rf_rec  <- rf_cm$byClass["Sensitivity"]
rf_f1   <- rf_cm$byClass["F1"]
rf_auc  <- as.numeric(auc(roc(test$Treatment_Cat, rf_probs)))

cat("\nBest config (5-fold CV): mtry =", best_cfg$mtry,
    ", maxnodes =", best_cfg$maxnodes, ", ntree =", best_cfg$ntree)
cat("\nRandom Forest OOB error: ", round(oob_err, 3))
cat("\nAccuracy: ", round(rf_acc, 3))
cat("\nPrecision:", round(rf_prec, 3))
cat("\nRecall:   ", round(rf_rec, 3))
cat("\nF1-Score: ", round(rf_f1, 3))
cat("\nROC-AUC:  ", round(rf_auc, 3), "\n")
```

```{r rf-importance-plot, fig.width=10, fig.height=5, fig.align='center'}
# Dual Variable Importance Plot (Accuracy & Gini)
imp_matrix <- importance(rf_model)
base_names <- sub('(Gender_Category|Family_History|Tech_Company_Cat).*', '\\1', rownames(imp_matrix))

acc_imp  <- rowsum(imp_matrix[, "MeanDecreaseAccuracy"], base_names, reorder = FALSE)[, 1]
gini_imp <- rowsum(imp_matrix[, "MeanDecreaseGini"], base_names, reorder = FALSE)[, 1]

rf_imp_full <- data.frame(
  Predictor = names(acc_imp),
  Accuracy  = round(acc_imp, 2),
  Gini      = round(gini_imp, 2)
)

par(mfrow = c(1, 2), mar = c(8.5, 4.5, 3.5, 1.5))

# Plot A: Accuracy
df_acc <- rf_imp_full %>% arrange(desc(Accuracy))
max_y_acc <- max(df_acc$Accuracy) * 1.15
bp1 <- barplot(df_acc$Accuracy, names.arg = df_acc$Predictor, col = "steelblue", 
               main = "Variable Importance\n(Mean Decrease Accuracy)",
               ylab = "Accuracy Decrease", ylim = c(min(0, min(df_acc$Accuracy) * 1.2), max_y_acc),
               las = 2, cex.names = 0.75, cex.main = 0.95)
text(bp1, ifelse(df_acc$Accuracy >= 0, df_acc$Accuracy + (max_y_acc * 0.02), df_acc$Accuracy - (max_y_acc * 0.04)),
     labels = df_acc$Accuracy, cex = 0.7, xpd = TRUE)

# Plot B: Gini
df_gini <- rf_imp_full %>% arrange(desc(Gini))
max_y_gini <- max(df_gini$Gini) * 1.15
bp2 <- barplot(df_gini$Gini, names.arg = df_gini$Predictor, col = "darkorange", 
               main = "Variable Importance\n(Mean Decrease Gini)",
               ylab = "Gini Impurity Decrease", ylim = c(0, max_y_gini),
               las = 2, cex.names = 0.75, cex.main = 0.95)
text(bp2, df_gini$Gini + (max_y_gini * 0.02), labels = df_gini$Gini, cex = 0.7, xpd = TRUE)

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

```{r}
comparison_rf <- tibble(
  Model = c("Binary Logistic Regression", "Random Forest"),
  Accuracy  = c(round(acc_val, 3), round(rf_acc, 3)),
  Precision = c(round(prec_val, 3), round(rf_prec, 3)),
  Recall    = c(round(rec_val, 3), round(rf_rec, 3)),
  `F1-Score` = c(round(logit_f1, 3), round(rf_f1, 3)),
  `ROC-AUC` = c(round(auc_value, 3), round(rf_auc, 3))
)
knitr::kable(comparison_rf, caption = "Test-Set Comparison: Logistic vs. Random Forest")
```

### Random Forest Benchmark Results

The Random Forest achieves an out-of-sample **accuracy of `r round(rf_acc * 100, 1)`% with a ROC-AUC of `r round(rf_auc, 3)`** (`r round(rf_f1, 3)` F1), compared with the logistic model's accuracy of `r round(acc_val * 100, 1)`% and AUC of `r round(auc_value, 3)`. The two classifiers perform very similarly, with the linear logistic model holding a slight edge on this small tabular sample.

Critically, the Random Forest's variable-importance ranking **corroborates the logistic finding**: `Family_History` is far and away the most influential predictor, while workplace measures, tech-company status, and gender contribute only marginal splits.

## 5.5 Principal Component Analysis (PCA) with Random Forest

We test empirically whether feeding the Random Forest PCA components---fitted strictly on the training set to prevent leakage---improves predictive performance.

```{r}
# Pure Train-Only PCA to eliminate Data Leakage
X_train_mat <- rf_X
X_test_mat  <- rf_X_test

pca_fit <- prcomp(X_train_mat, center = TRUE, scale. = TRUE)
cum_var <- cumsum(pca_fit$sdev^2 / sum(pca_fit$sdev^2))

Z_train <- predict(pca_fit, newdata = X_train_mat)
Z_test  <- predict(pca_fit, newdata = X_test_mat)

pca_summary <- tibble(
  Component = 1:length(cum_var),
  `Prop. Variance` = round(pca_fit$sdev^2 / sum(pca_fit$sdev^2), 3),
  `Cumulative`      = round(cum_var, 3)
)
knitr::kable(pca_summary, caption = "Variance explained by PCA components")
```

```{r}
# 5-fold CV for PCA Random Forest (k = 8)
k_pca <- 8

set.seed(2026)
pca_folds <- createFolds(train$Treatment_Cat, k = 5, list = TRUE, returnTrain = TRUE)

pca_grid <- expand.grid(
  mtry     = c(1, 2, 3),
  maxnodes = c(8, 16, 32),
  ntree    = c(100, 300)
)

pca_cv_results <- data.frame()
for (g in seq_len(nrow(pca_grid))) {
  aucs <- numeric(5)
  for (f in seq_along(pca_folds)) {
    tr_idx <- pca_folds[[f]]
    set.seed(123)
    rf_pca_f <- randomForest(
      x = Z_train[tr_idx, 1:k_pca, drop = FALSE], y = train$Treatment_Cat[tr_idx],
      mtry = pca_grid$mtry[g], maxnodes = pca_grid$maxnodes[g],
      ntree = pca_grid$ntree[g]
    )
    p_f <- predict(rf_pca_f, Z_train[-tr_idx, 1:k_pca, drop = FALSE], type = "prob")[, "Yes"]
    aucs[f] <- as.numeric(auc(roc(train$Treatment_Cat[-tr_idx], p_f)))
  }
  pca_cv_results <- rbind(pca_cv_results, data.frame(
    mtry = pca_grid$mtry[g], maxnodes = pca_grid$maxnodes[g],
    ntree = pca_grid$ntree[g], CV_AUC = round(mean(aucs), 4)
  ))
}
pca_best_cfg <- pca_cv_results[which.max(pca_cv_results$CV_AUC), ]

# Retrain best PCA model
set.seed(123)
rf_pca <- randomForest(
  x = Z_train[, 1:k_pca, drop = FALSE], y = train$Treatment_Cat,
  mtry = pca_best_cfg$mtry, maxnodes = pca_best_cfg$maxnodes,
  ntree = pca_best_cfg$ntree
)

p_pca_rf  <- predict(rf_pca, Z_test[, 1:k_pca, drop = FALSE], type = "prob")[, "Yes"]
cl_pca_rf <- factor(ifelse(p_pca_rf >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
cm_pca_rf <- confusionMatrix(cl_pca_rf, test$Treatment_Cat, positive = "Yes")

pca_rf_auc  <- as.numeric(auc(roc(test$Treatment_Cat, p_pca_rf)))
pca_rf_acc  <- cm_pca_rf$overall["Accuracy"]
pca_rf_prec <- cm_pca_rf$byClass["Precision"]
pca_rf_rec  <- cm_pca_rf$byClass["Sensitivity"]
pca_rf_f1   <- cm_pca_rf$byClass["F1"]

cat("\nBest PCA config (5-fold CV): mtry =", pca_best_cfg$mtry,
    ", maxnodes =", pca_best_cfg$maxnodes, ", ntree =", pca_best_cfg$ntree)
cat("\nPCA-RF (k=8) Accuracy:", round(pca_rf_acc, 3),
    " Precision:", round(pca_rf_prec, 3),
    " Recall:", round(pca_rf_rec, 3),
    " F1:", round(pca_rf_f1, 3),
    " ROC-AUC:", round(pca_rf_auc, 3), "\n")
```

```{r}
comparison_pca <- tibble(
  Model = c("Binary Logistic", "Random Forest (raw)", "Random Forest (PCA k=8)"),
  Accuracy  = c(round(acc_val, 3), round(rf_acc, 3), round(pca_rf_acc, 3)),
  Precision = c(round(prec_val, 3), round(rf_prec, 3), round(pca_rf_prec, 3)),
  Recall    = c(round(rec_val, 3), round(rf_rec, 3), round(pca_rf_rec, 3)),
  `F1-Score` = c(round(logit_f1, 3), round(rf_f1, 3), round(pca_rf_f1, 3)),
  `ROC-AUC` = c(round(auc_value, 3), round(rf_auc, 3), round(pca_rf_auc, 3))
)
knitr::kable(comparison_pca, caption = "Test-Set Comparison: Logistic vs. Random Forest (raw and PCA)")
```

```{r model-comparison-roc, fig.width=7, fig.height=6, fig.align='center'}
# 3-Model Overlay ROC Comparison Curve
roc_logit  <- roc(test$Treatment_Cat, test_probs)
roc_rf_raw <- roc(test$Treatment_Cat, rf_probs)
roc_rf_pca <- roc(test$Treatment_Cat, p_pca_rf)

par(mar = c(4.5, 4.5, 3, 1))
plot(roc_logit, col = "firebrick", lwd = 2, legacy.axes = TRUE, ylim = c(0, 1),
     main = "Test-Set ROC Curves Comparison across Models",
     xlab = "False Positive Rate (1 - Specificity)", 
     ylab = "True Positive Rate (Sensitivity)")

plot(roc_rf_raw, col = "forestgreen", lwd = 2, add = TRUE)
plot(roc_rf_pca, col = "dodgerblue", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "gray50")

legend("bottomright", 
       legend = c(
         paste0("Binary Logistic (AUC = ", round(auc_value, 3), ")"),
         paste0("Random Forest Raw (AUC = ", round(rf_auc, 3), ")"),
         paste0("Random Forest PCA k=8 (AUC = ", round(pca_rf_auc, 3), ")")
       ),
       col = c("firebrick", "forestgreen", "dodgerblue"), lwd = 2, bty = "n", cex = 0.85)

par(mar = c(5.1, 4.1, 4.1, 2.1))
```

### PCA-RF Benchmark Results

With $k=8$ PCA components retaining $\approx$ `r round(cum_var[8] * 100, 1)`% of variance, the PCA-transformed Random Forest achieves a **ROC-AUC of `r round(pca_rf_auc, 3)` and accuracy of `r round(pca_rf_acc * 100, 1)`%** on the test set. On ROC-AUC, it slightly edges out the raw Random Forest and matches the linear baseline. The practical conclusion remains robust: personal family history dominates the decision to seek treatment, unaffected by corporate environment variables or model architecture choice.