---
title: 'Activity 2 — Task 5: Binary Logistic Regression Modeling'
author: "Group 08"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 5 Binary Logistic Regression Modeling

## 5.1 Research Question and Data Preparation

This final task addresses the third research question for Activity 2: **Which personal factors (age, gender, family history) and environmental factors (employer prioritization, coworker reactions) increase or decrease the probability of an employee seeking professional mental health treatment? How accurately can a model classify those who need treatment?**

Because the target variable (`Treatment_Cat`) is binary (Yes/No), we employ a Binary Logistic Regression model.

To maintain strict comparability with the continuous regression analysis, we utilize the exact same globally randomized 80:20 split (seed 2026).

```{r}
library(tidyverse)
library(caret)
library(pROC)
library(broom)

data <- read.csv("dataset/osmi_clean_combined.csv")

# Ensure reference levels are set logically for interpretation
data <- data %>%
  mutate(
    Treatment_Cat = factor(Treatment_Cat, levels = c("No", "Yes")),
    Gender_Category = relevel(factor(Gender_Category), ref = "Male"),
    Family_History = relevel(factor(Family_History), ref = "No"),
    Tech_Company_Cat = relevel(factor(Tech_Company_Cat), ref = "No")
  )

set.seed(2026)
train_idx <- sample(nrow(data), floor(0.8 * nrow(data)))
train <- data[train_idx, ]
test  <- data[-train_idx, ]

cat("Training cases:", nrow(train), "| Test cases:", nrow(test), "\n")
```

## 5.2 Model Fitting and Odds Ratios

We fit the logistic regression model on the training set using both demographic and corporate environment predictors.

```{r}
logit_model <- glm(
  Treatment_Cat ~ Age + Gender_Category + Family_History + 
    Employer_Mental_Importance + Employer_Phys_Importance + 
    Coworker_Reaction_Score + Tech_Company_Cat + as.factor(Year),
  family = binomial(link = "logit"),
  data = train
)

# Extract Odds Ratios (OR) and Confidence Intervals
model_matrix <- summary(logit_model)$coefficients
wald_ci <- confint.default(logit_model)

odds_ratios <- tibble(
  Predictor = rownames(model_matrix),
  Estimate = model_matrix[, "Estimate"],
  `Odds Ratio (OR)` = exp(Estimate),
  `P-value` = model_matrix[, "Pr(>|z|)"],
  Significant = `P-value` < 0.05
) %>%
  filter(Predictor != "(Intercept)") %>%
  mutate(
    across(c(Estimate, `Odds Ratio (OR)`), ~ round(.x, 3)),
    `P-value` = format.pval(`P-value`, digits = 3, eps = 0.001)
  )

knitr::kable(odds_ratios, caption = "Adjusted Odds Ratios for Seeking Treatment")
```

### What drives the decision to seek treatment?

By examining the **Odds Ratios (OR)**, where $OR > 1$ indicates increased probability and $OR < 1$ indicates decreased probability, we uncover a fascinating divide between personal and environmental factors:

- **Personal Family History Dominates:** The strongest predictor by a massive margin is personal biology. Employees with a known family history of mental illness have incredibly high odds of seeking treatment compared to those without. Age also shows a slight, statistically significant positive correlation ($OR = 1.024$, $p = 0.013$). Gender shows a smaller but still significant association: Female employees have higher odds of seeking treatment than Male employees ($OR = 1.644$, $p = 0.008$), while the Other category does not differ from Male.

- **Environmental Factors are Weak:** Surprisingly, workplace variables have almost zero impact on the individual's decision to seek clinical help. The importance the employer places on mental/physical health, the tech-company status, and even the coworker reaction score all yielded non-significant p-values.

In short: **The decision to seek professional treatment is driven primarily by family history, with smaller contributions from age and female gender, and essentially none from corporate policy.**

## 5.3 Out-of-Sample Classification Performance

To answer the second half of the research question ("How accurately can the model classify?"), we apply the fitted model to the 211 unseen records in the test set.

```{r}
# Generate probabilities and classes on the Test Set
test_probs <- predict(logit_model, newdata = test, type = "response")
test_preds <- factor(ifelse(test_probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
actual_class <- test$Treatment_Cat

# Confusion Matrix
conf_matrix <- confusionMatrix(test_preds, actual_class, positive = "Yes")
print(conf_matrix$table)

# Extract Metrics for inline text
acc_val <- conf_matrix$overall["Accuracy"]
prec_val <- conf_matrix$byClass["Pos Pred Value"]
rec_val <- conf_matrix$byClass["Sensitivity"]

# ROC and AUC
roc_curve <- roc(actual_class, test_probs)
auc_value <- as.numeric(auc(roc_curve))

cat("\nAccuracy: ", round(acc_val, 3))
cat("\nPrecision:", round(prec_val, 3))
cat("\nRecall:   ", round(rec_val, 3))
cat("\nROC-AUC:  ", round(auc_value, 3), "\n")

plot(roc_curve, col = "firebrick", lwd = 2, 
     main = paste("ROC Curve (Test Set) - AUC =", round(auc_value, 3)))
```

### Conclusion for Research Question 3

The model successfully identifies whether an employee will seek professional mental health treatment with an overall **Accuracy of `r round(acc_val * 100, 1)`%** on completely unseen data.

More importantly, the **ROC-AUC score is `r round(auc_value, 3)`**. This means that if we randomly select one employee who sought treatment and one who did not, the model has an `r round(auc_value * 100, 1)`% probability of correctly ranking the treatment-seeker with a higher risk score.

The precision (`r round(prec_val, 3)`) and recall (`r round(rec_val, 3)`) demonstrate a well-balanced classifier that doesn't just guess "Yes" for everyone.

**Final Answer:** We can classify treatment-seekers with high accuracy ($\approx$ `r round(acc_val * 100, 1)`%). The analysis proves that the decision to seek treatment is profoundly driven by **personal biology** (specifically family history), with smaller positive associations for **age and female gender**, while corporate environmental factors (like supportive coworkers or HR policies) fail to statistically alter an employee's underlying probability of seeking professional medical care.

## 5.4 Random Forest Benchmark

The logistic model is linear and assumes log-odds are a linear function of the predictors. To check whether a *nonlinear* tree ensemble can capture additional structure---interactions between predictors or curvilinear effects---we fit a **Random Forest** classifier on the exact same binary target (`Treatment_Cat`) and the same training/test partition (seed 2026, 844/211). Random Forest aggregates many decision trees, each trained on a bootstrap sample with a random subset of predictors, which lets it model nonlinearity and interactions without parametric assumptions.

```{r}
# Random Forest needs its own package
library(randomForest)

# Numeric design matrix (same predictors as the logistic model) for x/y interface
rf_X <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = train
)[, -1]
rf_y <- train$Treatment_Cat

# ---- 5-fold stratified cross-validation to tune the random forest ----
set.seed(2026)
folds <- createFolds(rf_y, k = 5, list = TRUE, returnTrain = TRUE)

grid <- expand.grid(
  mtry     = c(2, 3, 5),
  maxnodes = c(8, 16, 32),
  ntree    = c(100, 300)
)

cv_results <- data.frame()
for (g in seq_len(nrow(grid))) {
  aucs <- numeric(5)
  for (f in seq_along(folds)) {
    tr_idx <- folds[[f]]
    set.seed(123)
    rf_f <- randomForest(
      x = rf_X[tr_idx, , drop = FALSE], y = rf_y[tr_idx],
      mtry = grid$mtry[g], maxnodes = grid$maxnodes[g], ntree = grid$ntree[g]
    )
    p_f <- predict(rf_f, rf_X[-tr_idx, , drop = FALSE], type = "prob")[, "Yes"]
    aucs[f] <- as.numeric(auc(roc(rf_y[-tr_idx], p_f)))
  }
  cv_results <- rbind(cv_results, data.frame(
    mtry = grid$mtry[g], maxnodes = grid$maxnodes[g],
    ntree = grid$ntree[g], CV_AUC = round(mean(aucs), 4)
  ))
}
best_cfg <- cv_results[which.max(cv_results$CV_AUC), ]
knitr::kable(cv_results[order(-cv_results$CV_AUC), ],
             caption = "5-fold CV tuning of the random forest (mean validation AUC)")

# ---- Retrain best config on the full training set and score the test set ----
set.seed(123)
rf_model <- randomForest(
  x = rf_X, y = rf_y,
  mtry = best_cfg$mtry, maxnodes = best_cfg$maxnodes,
  ntree = best_cfg$ntree, importance = TRUE
)

# Out-of-bag error rate
oob_err <- mean(rf_model$err.rate[, 1])

# Predict on the held-out test set (use the same model.matrix structure)
rf_X_test <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = test
)[, -1]
rf_probs <- predict(rf_model, rf_X_test, type = "prob")[, "Yes"]
rf_preds <- factor(ifelse(rf_probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))

rf_cm <- confusionMatrix(rf_preds, test$Treatment_Cat, positive = "Yes")
print(rf_cm$table)

# Metrics
rf_acc  <- rf_cm$overall["Accuracy"]
rf_prec <- rf_cm$byClass["Precision"]
rf_rec  <- rf_cm$byClass["Sensitivity"]
rf_f1   <- rf_cm$byClass["F1"]
rf_auc  <- as.numeric(auc(roc(test$Treatment_Cat, rf_probs)))

cat("\nBest config (5-fold CV): mtry =", best_cfg$mtry,
    ", maxnodes =", best_cfg$maxnodes, ", ntree =", best_cfg$ntree)
cat("\nRandom Forest OOB error: ", round(oob_err, 3))
cat("\nAccuracy: ", round(rf_acc, 3))
cat("\nPrecision:", round(rf_prec, 3))
cat("\nRecall:   ", round(rf_rec, 3))
cat("\nF1-Score: ", round(rf_f1, 3))
cat("\nROC-AUC:  ", round(rf_auc, 3), "\n")
```

```{r}
# Variable importance: how much each predictor reduces classification accuracy.
# Aggregate dummy-level rows back to their original predictors.
raw_imp <- importance(rf_model)[, "MeanDecreaseAccuracy", drop = FALSE]
base_names <- sub('(Gender_Category|Family_History|Tech_Company_Cat).*', '\\1',
                  rownames(raw_imp))
imp_by_predictor <- rowsum(raw_imp[, 1], base_names, reorder = FALSE)
rf_imp <- imp_by_predictor[, 1]
names(rf_imp) <- rownames(imp_by_predictor)

rf_imp_df <- data.frame(
  Predictor = names(rf_imp),
  MeanDecreaseAccuracy = round(rf_imp, 2)
) %>%
  arrange(desc(MeanDecreaseAccuracy))
knitr::kable(rf_imp_df, caption = "Random Forest Variable Importance (Mean Decrease in Accuracy)")
```

```{r}
# Comparison of the two classifiers on the held-out test set
logit_f1 <- 2 * prec_val * rec_val / (prec_val + rec_val)
comparison_rf <- tibble(
  Model = c("Binary Logistic Regression", "Random Forest"),
  Accuracy  = c(round(acc_val, 3), round(rf_acc, 3)),
  Precision = c(round(prec_val, 3), round(rf_prec, 3)),
  Recall    = c(round(rec_val, 3), round(rf_rec, 3)),
  `F1-Score` = c(round(logit_f1, 3), round(rf_f1, 3)),
  `ROC-AUC` = c(round(auc_value, 3), round(rf_auc, 3))
)
knitr::kable(comparison_rf, caption = "Test-Set Comparison: Logistic vs. Random Forest")
```

### Random Forest Benchmark Results

The Random Forest (hyperparameters chosen by 5-fold cross-validation: `mtry = `r best_cfg$mtry`, `maxnodes = `r best_cfg$maxnodes`, `ntree = `r best_cfg$ntree`) achieves an out-of-sample **accuracy of `r round(rf_acc * 100, 1)`% with a ROC-AUC of `r round(rf_auc, 3)`** (`r round(rf_f1, 3)` F1), compared with the logistic model's accuracy of `r round(acc_val * 100, 1)`% and AUC of `r round(auc_value, 3)`. The two classifiers perform very similarly, with the linear logistic model holding a slight edge on this small tabular sample---consistent with the expectation that a flexible tree ensemble has little extra signal to extract from only 844 training rows.

Critically, the Random Forest's variable-importance ranking **corroborates the logistic finding**:            `Family_History` is far and away the most influential predictor (mean decrease in accuracy $\approx$ `r round(rf_imp["Family_History"], 0)`), while the workplace measures, tech-company status, and gender contribute only marginal splits. Whether modeled linearly or nonlinearly, family history dominates the decision to seek treatment. The one nuance is that the linear logistic model detects a modest female-versus-male odds advantage ($OR=1.644$) that the partitions of this small tree ensemble do not exploit.

## 5.5 Principal Component Analysis (PCA) with Random Forest

The predictors include several correlated workplace ratings and demographic measures spanning different scales (Age 19--65, ratings 0--10, binary indicators). **Principal Component Analysis (PCA)** decorrelates these features and compresses them into a smaller set of orthogonal components ordered by the variance they explain. This is a common preprocessing step for reducing multicollinearity and dimensionality. For a Random Forest, we test empirically whether feeding the model PCA components---rather than the original correlated features---changes predictive performance.

The transformation is fitted **only on the training set** (standardizing each feature and extracting the principal-component loadings) and then applied to the test set, avoiding data leakage.

```{r}
# Build the numeric design matrix (same predictors as the logistic model);
# one-hot encode the categorical predictors, dropping base levels as in model.matrix
X_all <- model.matrix(
  ~ Age + Gender_Category + Family_History + Employer_Mental_Importance +
    Employer_Phys_Importance + Coworker_Reaction_Score + Tech_Company_Cat,
  data = rbind(train, test)
)[, -1]

n_train <- nrow(train)
X_train_mat <- X_all[1:n_train, ]
X_test_mat  <- X_all[(n_train + 1):nrow(X_all), ]

# PCA on the TRAINING set only (center + scale), then transform test
pca_fit <- prcomp(X_train_mat, center = TRUE, scale. = TRUE)

# Proportion of variance explained by the first k components
cum_var <- cumsum(pca_fit$sdev^2 / sum(pca_fit$sdev^2))

# Transform both sets using the training PCA loadings
Z_train <- predict(pca_fit, newdata = X_train_mat)
Z_test  <- predict(pca_fit, newdata = X_test_mat)

# Variance explained
pca_summary <- tibble(
  Component = 1:length(cum_var),
  `Prop. Variance` = round(pca_fit$sdev^2 / sum(pca_fit$sdev^2), 3),
  `Cumulative`      = round(cum_var, 3)
)
knitr::kable(pca_summary, caption = "Variance explained by PCA components")
```

```{r}
# ---- 5-fold stratified CV to tune the PCA random forest (k = 8 components) ----
k_pca <- 8   # drop the lowest-variance component from the 9-feature design matrix

set.seed(2026)
pca_folds <- createFolds(train$Treatment_Cat, k = 5, list = TRUE, returnTrain = TRUE)

pca_grid <- expand.grid(
  mtry     = c(1, 2, 3),
  maxnodes = c(8, 16, 32),
  ntree    = c(100, 300)
)

pca_cv_results <- data.frame()
for (g in seq_len(nrow(pca_grid))) {
  aucs <- numeric(5)
  for (f in seq_along(pca_folds)) {
    tr_idx <- pca_folds[[f]]
    set.seed(123)
    rf_pca_f <- randomForest(
      x = Z_train[tr_idx, 1:k_pca, drop = FALSE], y = train$Treatment_Cat[tr_idx],
      mtry = pca_grid$mtry[g], maxnodes = pca_grid$maxnodes[g],
      ntree = pca_grid$ntree[g]
    )
    p_f <- predict(rf_pca_f, Z_train[-tr_idx, 1:k_pca, drop = FALSE], type = "prob")[, "Yes"]
    aucs[f] <- as.numeric(auc(roc(train$Treatment_Cat[-tr_idx], p_f)))
  }
  pca_cv_results <- rbind(pca_cv_results, data.frame(
    mtry = pca_grid$mtry[g], maxnodes = pca_grid$maxnodes[g],
    ntree = pca_grid$ntree[g], CV_AUC = round(mean(aucs), 4)
  ))
}
pca_best_cfg <- pca_cv_results[which.max(pca_cv_results$CV_AUC), ]
knitr::kable(pca_cv_results[order(-pca_cv_results$CV_AUC), ],
             caption = "5-fold CV tuning of the PCA random forest (k = 8 components, mean validation AUC)")

# ---- Retrain best PCA config on the full training set, score the test set ----
set.seed(123)
rf_pca <- randomForest(
  x = Z_train[, 1:k_pca, drop = FALSE], y = train$Treatment_Cat,
  mtry = pca_best_cfg$mtry, maxnodes = pca_best_cfg$maxnodes,
  ntree = pca_best_cfg$ntree
)

p_pca_rf <- predict(rf_pca, Z_test[, 1:k_pca, drop = FALSE], type = "prob")[, "Yes"]
cl_pca_rf <- factor(ifelse(p_pca_rf >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
cm_pca_rf <- confusionMatrix(cl_pca_rf, test$Treatment_Cat, positive = "Yes")

pca_rf_auc  <- as.numeric(auc(roc(test$Treatment_Cat, p_pca_rf)))
pca_rf_acc  <- cm_pca_rf$overall["Accuracy"]
pca_rf_prec <- cm_pca_rf$byClass["Precision"]
pca_rf_rec  <- cm_pca_rf$byClass["Sensitivity"]
pca_rf_f1   <- cm_pca_rf$byClass["F1"]

cat("\nBest PCA config (5-fold CV): mtry =", pca_best_cfg$mtry,
    ", maxnodes =", pca_best_cfg$maxnodes, ", ntree =", pca_best_cfg$ntree)
cat("\nPCA-RF (k=8) Accuracy:", round(pca_rf_acc, 3),
    " Precision:", round(pca_rf_prec, 3),
    " Recall:", round(pca_rf_rec, 3),
    " F1:", round(pca_rf_f1, 3),
    " ROC-AUC:", round(pca_rf_auc, 3), "\n")
```

```{r}
# 3-way comparison: logistic, raw random forest, and PCA random forest
comparison_pca <- tibble(
  Model = c("Binary Logistic", "Random Forest (raw)", "Random Forest (PCA k=8)"),
  Accuracy  = c(round(acc_val, 3), round(rf_acc, 3), round(pca_rf_acc, 3)),
  Precision = c(round(prec_val, 3), round(rf_prec, 3), round(pca_rf_prec, 3)),
  Recall    = c(round(rec_val, 3), round(rf_rec, 3), round(pca_rf_rec, 3)),
  `F1-Score` = c(round(logit_f1, 3), round(rf_f1, 3), round(pca_rf_f1, 3)),
  `ROC-AUC` = c(round(auc_value, 3), round(rf_auc, 3), round(pca_rf_auc, 3))
)
knitr::kable(comparison_pca, caption = "Test-Set Comparison: Logistic vs. Random Forest (raw and PCA)")
```

### PCA-RF Benchmark Results

With $k=8$ PCA components (which, from the nine-predictor design matrix, retain $\approx$ `r round(cum_var[8] * 100, 1)`% of the variance by dropping the single lowest-variance component) and the same 5-fold cross-validation used for the raw forest---here selecting `mtry = `r pca_best_cfg$mtry`, `maxnodes = `r pca_best_cfg$maxnodes`, `ntree = `r pca_best_cfg$ntree`---the PCA-transformed Random Forest attains a **ROC-AUC of `r round(pca_rf_auc, 3)` and accuracy of `r round(pca_rf_acc * 100, 1)`%** on the held-out test set. On ROC-AUC it edges out both the raw-feature Random Forest and the linear logistic model (`r round(auc_value, 3)`); on accuracy it lands just below the logistic baseline. The modest gain and near-matching accuracy reflect the fact that the treatment decision is driven by one dominant binary predictor (family history) that a linear model already captures, so even a tuned nonlinear ensemble and a mild dimensionality reduction add little beyond ranking better on borderline cases. The practical takeaway is unchanged: no PCA preprocessing alters the substantive conclusion that treatment-seeking is governed by personal biology rather than the corporate environment or gender.
