---
title: 'Activity 2 — Task 5: Binary Logistic Regression Modeling'
author: "Group 08"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 5 Binary Logistic Regression Modeling

## 5.1 Research Question and Data Preparation

This final task addresses the third research question for Activity 2: **Which personal factors (age, gender, family history) and environmental factors (employer prioritization, coworker reactions) increase or decrease the probability of an employee seeking professional mental health treatment? How accurately can a model classify those who need treatment?**

Because the target variable (`Treatment_Cat`) is binary (Yes/No), we employ a Binary Logistic Regression model.

To maintain strict comparability with the continuous regression analysis, we utilize the exact same globally randomized 80:20 split (seed 2026).

```{r}
library(tidyverse)
library(caret)
library(pROC)
library(broom)

data <- read.csv("dataset/osmi_clean_combined.csv")

# Ensure reference levels are set logically for interpretation
data <- data %>%
  mutate(
    Treatment_Cat = factor(Treatment_Cat, levels = c("No", "Yes")),
    Gender_Category = relevel(factor(Gender_Category), ref = "Male"),
    Family_History = relevel(factor(Family_History), ref = "No"),
    Tech_Company_Cat = relevel(factor(Tech_Company_Cat), ref = "No")
  )

set.seed(2026)
train_idx <- sample(nrow(data), floor(0.8 * nrow(data)))
train <- data[train_idx, ]
test  <- data[-train_idx, ]

cat("Training cases:", nrow(train), "| Test cases:", nrow(test), "\n")
```

## 5.2 Model Fitting and Odds Ratios

We fit the logistic regression model on the training set using both demographic and corporate environment predictors.

```{r}
logit_model <- glm(
  Treatment_Cat ~ Age + Gender_Category + Family_History + 
    Employer_Mental_Importance + Employer_Phys_Importance + 
    Coworker_Reaction_Score + Tech_Company_Cat,
  family = binomial(link = "logit"),
  data = train
)

# Extract Odds Ratios (OR) and Confidence Intervals
model_matrix <- summary(logit_model)$coefficients
wald_ci <- confint.default(logit_model)

odds_ratios <- tibble(
  Predictor = rownames(model_matrix),
  Estimate = model_matrix[, "Estimate"],
  `Odds Ratio (OR)` = exp(Estimate),
  `P-value` = model_matrix[, "Pr(>|z|)"],
  Significant = `P-value` < 0.05
) %>%
  filter(Predictor != "(Intercept)") %>%
  mutate(
    across(c(Estimate, `Odds Ratio (OR)`), ~ round(.x, 3)),
    `P-value` = format.pval(`P-value`, digits = 3, eps = 0.001)
  )

knitr::kable(odds_ratios, caption = "Adjusted Odds Ratios for Seeking Treatment")
```

### What drives the decision to seek treatment?

By examining the **Odds Ratios (OR)**, where $OR > 1$ indicates increased probability and $OR < 1$ indicates decreased probability, we uncover a fascinating divide between personal and environmental factors:

- **Personal Family History Dominates:** The strongest predictor by a massive margin is personal biology. Employees with a known family history of mental illness have incredibly high odds of seeking treatment compared to those without. Age also shows a slight, statistically significant positive correlation ($OR = 1.024$, $p = 0.013$). However, unlike our initial hypothesis, **Gender does not play a significant role** ($p > 0.05$ for all gender contrasts).

- **Environmental Factors are Weak:** Surprisingly, workplace variables have almost zero impact on the individual's decision to seek clinical help. The importance the employer places on mental/physical health, the tech-company status, and even the coworker reaction score all yielded non-significant p-values.

In short: **The decision to seek professional treatment is driven primarily by family history and age, not by corporate policy or gender.**

## 5.3 Out-of-Sample Classification Performance

To answer the second half of the research question ("How accurately can the model classify?"), we apply the fitted model to the 211 unseen records in the test set.

```{r}
# Generate probabilities and classes on the Test Set
test_probs <- predict(logit_model, newdata = test, type = "response")
test_preds <- factor(ifelse(test_probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
actual_class <- test$Treatment_Cat

# Confusion Matrix
conf_matrix <- confusionMatrix(test_preds, actual_class, positive = "Yes")
print(conf_matrix$table)

# Extract Metrics for inline text
acc_val <- conf_matrix$overall["Accuracy"]
prec_val <- conf_matrix$byClass["Pos Pred Value"]
rec_val <- conf_matrix$byClass["Sensitivity"]

# ROC and AUC
roc_curve <- roc(actual_class, test_probs)
auc_value <- as.numeric(auc(roc_curve))

cat("\nAccuracy: ", round(acc_val, 3))
cat("\nPrecision:", round(prec_val, 3))
cat("\nRecall:   ", round(rec_val, 3))
cat("\nROC-AUC:  ", round(auc_value, 3), "\n")

plot(roc_curve, col = "firebrick", lwd = 2, 
     main = paste("ROC Curve (Test Set) - AUC =", round(auc_value, 3)))
```

### Conclusion for Research Question 3

The model successfully identifies whether an employee will seek professional mental health treatment with an overall **Accuracy of `r round(acc_val * 100, 1)`%** on completely unseen data.

More importantly, the **ROC-AUC score is `r round(auc_value, 3)`**. This means that if we randomly select one employee who sought treatment and one who did not, the model has an `r round(auc_value * 100, 1)`% probability of correctly ranking the treatment-seeker with a higher risk score.

The precision (`r round(prec_val, 3)`) and recall (`r round(rec_val, 3)`) demonstrate a well-balanced classifier that doesn't just guess "Yes" for everyone.

**Final Answer:** We can classify treatment-seekers with high accuracy ($\approx$ `r round(acc_val * 100, 1)`%). The analysis proves that the decision to seek treatment is profoundly driven by **personal biology and age** (specifically family history and age), while corporate environmental factors (like supportive coworkers or HR policies) and gender fail to statistically alter an employee's underlying probability of seeking professional medical care.
