---
title: "Activity 1 — Task 1: Data Preprocessing"
author: "Group 08"
date: "2026-08-16"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = TRUE)
```

## Read Data

We start by loading the raw dataset `ds_salaries.csv` (607 observations, 11
variables) and inspecting its structure.

```{r}
library(tidyverse)
data <- read.csv("dataset/ds_salaries.csv")
head(data)
summary(data)
```

# 1 Data Preprocessing

## 1.1 Remove the ID column

The first column `X` is only a row index (0–606). It carries no information,
so we remove it to avoid confusing the models later.

```{r}
data$X <- NULL
head(data)
```

## 1.2 Check for missing values and empty strings

```{r}
missing_values <- colSums(is.na(data))
print(missing_values) ### The dataset has no Na values
```

```{r}
empty_strings <- colSums(data == "")
print(empty_strings) ### The dataset has no empty strings
```

The dataset contains **no missing values** and **no empty strings**, so no
imputation or deletion is needed.

## 1.3 Remove duplicate records

```{r}
duplicate_rows <- data[duplicated(data), ]
print(nrow(duplicate_rows)) ### 42 dupplicated rows
```

```{r}
data <- data %>% distinct()
cat("Rows after removing duplicates:", nrow(data), "\n") ### 607 -> 565
```

42 rows are exact duplicates of other rows. We keep the first occurrence of
each and remove the rest (**607 → 565**). Duplicates add no information and
would double-count observations in every later statistic (means, ANOVA group
sizes, model fitting).

## 1.4 Detect and handle outliers in `salary_in_usd`

### Boxplot of the raw salary

```{r}
ggplot(data, aes(x = "", y = salary_in_usd)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red", outlier.shape = 16) +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal() +
  labs(title = "Detecting Outliers in Salary (USD)", x = "", y = "Salary in USD")
```

### IQR rule (global)

```{r}
Q1 <- quantile(data$salary_in_usd, 0.25)
Q3 <- quantile(data$salary_in_usd, 0.75)
IQR_value <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

outlier_rows <- data %>%
  filter(salary_in_usd < lower_bound | salary_in_usd > upper_bound)

cat("Total IQR Outliers Detected:", nrow(outlier_rows), "\n") # 10 outliers detected
```

The IQR rule flags **10 observations** above 280,911 USD; none fall below the
lower bound.

### Decision: keep the outliers and log-transform

```{r}
data$log_salary <- log(data$salary_in_usd) ### natural logarithm (ln)
```

We **keep all 10 rows**. They are legitimate executive/director salaries
(EX/SE level, mostly US employers), not data-entry errors. Removing ~1.6% of an
already small sample (565) would bias group means and group sizes, and
winsorization would fabricate values. Instead, the skewness is handled at the
model level by adding `log_salary = ln(salary_in_usd)`, which later tasks use
as the response.

### Cross-check: per-group IQR within experience level

A single global IQR ignores the fact that salary naturally varies by
experience level. To verify that our flags are just high earners within each
group (not data errors), we repeat the IQR rule **within each experience
level**:

```{r}
group_outliers <- data %>%
  group_by(experience_level) %>%
  mutate(gQ1 = quantile(salary_in_usd, 0.25),
         gQ3 = quantile(salary_in_usd, 0.75),
         gIQR = gQ3 - gQ1,
         is_group_outlier = salary_in_usd < (gQ1 - 1.5 * gIQR) |
                            salary_in_usd > (gQ3 + 1.5 * gIQR)) %>%
  ungroup()
print(table(group_outliers$experience_level, group_outliers$is_group_outlier))
```

The flags are spread across all four levels (EN 2, MI 5, SE 3, EX 2), which
confirms they are simply the top earners of each group — supporting our keep
decision.

## 1.5 Group granular job titles into major categories

The dataset has 52 distinct job titles, too sparse for modeling. We group them
into **7 major categories** using priority-ordered keyword matching.
Management titles contain words like "Data Science" (e.g., "Head of Data
Science"), so they must be matched **first**; `Research Scientist` must be
matched before the generic `Scientist` pattern.

```{r}
data <- data %>%
  mutate(job_category = case_when(
    grepl("Director|Head|Manager|Lead", job_title, ignore.case = TRUE) ~ "Management_Lead",
    grepl("Machine Learning|ML Engineer|ML Scientist|AI|Computer Vision|NLP|Deep Learning|Researcher",
          job_title, ignore.case = TRUE) ~ "ML_AI_Engineer",
    grepl("Research Scientist", job_title, ignore.case = TRUE) ~ "Research",
    grepl("Data Engineer|Big Data|ETL|Cloud Data|Data Architect|Data Specialist|Data Science Engineer",
          job_title, ignore.case = TRUE) ~ "Data_Engineer",
    grepl("Data Scientist|Scientist", job_title, ignore.case = TRUE) ~ "Data_Scientist",
    grepl("Data Analyst", job_title, ignore.case = TRUE) ~ "Data_Analyst",
    TRUE ~ "Other"
  ))

stopifnot(sum(is.na(data$job_category)) == 0)

data$job_category <- factor(data$job_category)
print(table(data$job_category))
```

Every title maps to exactly one category (the `stopifnot` assertion passes),
and no title is left unmapped.

## 1.6 Encode categorical variables

### Ordinal encoding

`experience_level` and `company_size` have a natural order, so we encode them
as ordinal integers. The raw character columns are kept alongside for the
Factorial ANOVA in Task 3.

```{r}
data <- data %>%
  mutate(exp_level_ord = case_when(
    experience_level == "EN" ~ 1L,
    experience_level == "MI" ~ 2L,
    experience_level == "SE" ~ 3L,
    experience_level == "EX" ~ 4L
  ))

data <- data %>%
  mutate(company_size_ord = case_when(
    company_size == "S" ~ 1L,
    company_size == "M" ~ 2L,
    company_size == "L" ~ 3L
  ))
```

### One-hot encoding with explicit reference levels

We use one-hot encoding (drop the reference level) for the nominal variables.
The reference is set **explicitly** — never left to `factor()`'s alphabetical
order, which would silently pick e.g. `CT` instead of `FT` for
`employment_type`.

```{r}
one_hot <- function(df, col, prefix, ref_level) {
  f <- relevel(factor(df[[col]]), ref = ref_level)
  mm <- as.data.frame(model.matrix(~ f - 1, data = df)[, -1, drop = FALSE])
  colnames(mm) <- paste0(prefix, gsub("[^[:alnum:]]", "_", levels(f)[-1]))
  mm
}
```

`company_location` has 53 countries, 36 of which have at most 3 rows. Keeping
all of them would create ~52 near-empty dummies and unstable estimates, so we
collapse to the **Top 7 countries + Other** (US alone is ~63% of the sample).

```{r}
top_countries <- c("US", "GB", "CA", "DE", "IN", "FR", "ES")
data <- data %>%
  mutate(location_group = ifelse(company_location %in% top_countries,
                                 company_location, "Other"),
         residence_group = ifelse(employee_residence %in% top_countries,
                                  employee_residence, "Other"))

data <- bind_cols(
  data,
  one_hot(data, "job_category", "job_cat_", ref_level = "Data_Analyst"),
  one_hot(data, "employment_type", "empl_", ref_level = "FT"),
  one_hot(data, "location_group", "loc_", ref_level = "Other"),
  one_hot(data, "residence_group", "res_", ref_level = "Other")
)
```

`employment_type` is highly imbalanced (FT dominates; PT/CT/FL are rare). We
keep all four levels and one-hot encode them; the imbalance is acknowledged in
the report, and the ANOVA will use Type III sums of squares because of it.

For non-remote employees, the company location equals the employee residence,
so the two dummy sets would be strongly correlated in a regression. We add a
single binary `is_local` indicator as a low-collinearity alternative.

```{r}
data$is_local <- as.integer(data$company_location == data$employee_residence)
```

## 1.7 Scale numerical features

Numerical predictors are standardized (z-score) because penalized models
(Ridge/LASSO) require scale-invariant features. Raw columns are preserved for
interpretable models (MLR, ANOVA).

```{r}
data$salary_in_usd_scaled <- as.numeric(scale(data$salary_in_usd))
data$log_salary_scaled <- as.numeric(scale(data$log_salary))
data$remote_ratio_scaled <- as.numeric(scale(data$remote_ratio))
data$exp_level_ord_scaled <- as.numeric(scale(data$exp_level_ord))
data$company_size_ord_scaled <- as.numeric(scale(data$company_size_ord))
```

## 1.8 Final verification

```{r}
cat("Rows after cleaning:", nrow(data), "\n")                 ### 565
cat("Missing values after cleaning:", sum(is.na(data)), "\n") ### 0
cat("Duplicates after cleaning:", sum(duplicated(data)), "\n")### 0
```

### Contingency tables for downstream planning

These tables are printed for Members B & C: the first checks whether the
Job Category × Experience Level interaction has sparse cells; the second
quantifies the overlap between the location and residence dummies.

```{r}
print(table(data$job_category, data$experience_level))
```

```{r}
print(table(data$location_group, data$residence_group))
cat("is_local = 1 (location == residence):", sum(data$is_local),
    "of", nrow(data), "rows\n")
```

## 1.9 Data dictionary

```{r}
data_dictionary <- data.frame(
  Variable = c("work_year", "experience_level", "employment_type", "job_title",
               "salary", "salary_currency", "salary_in_usd", "employee_residence",
               "remote_ratio", "company_location", "company_size", "log_salary",
               "job_category", "exp_level_ord", "company_size_ord",
               "location_group", "residence_group", "is_local", "job_cat_*",
               "empl_*", "loc_*", "res_*", "*_scaled"),
  Type = c("numeric", "character", "character", "character", "numeric",
           "character", "numeric", "character", "numeric", "character",
           "character", "numeric", "factor", "ordinal (1-4)", "ordinal (1-3)",
           "factor", "factor", "binary (0/1)", "one-hot (6)", "one-hot (3)",
           "one-hot (7)", "one-hot (7)", "z-score"),
  Meaning = c("Year the salary was paid", "Experience level (EN/MI/SE/EX)",
              "Employment type (PT/FT/CT/FL)", "Original granular job title",
              "Gross salary in original currency", "ISO 4217 currency code",
              "Salary converted to USD (response)", "Employee's country of residence",
              "Remote work ratio (0/50/100)", "Employer's country",
              "Company size (S/M/L)", "ln(salary_in_usd) for modeling",
              "Grouped job category (7 levels)", "Ordinal encoding of experience",
              "Ordinal encoding of company size", "Top-7 countries + Other",
              "Top-7 countries + Other (residence)",
              "1 if company_location == employee_residence",
              "Dummies of job_category (ref: Data_Analyst)",
              "Dummies of employment_type (ref: FT)", "Dummies of location_group (ref: Other)",
              "Dummies of residence_group (ref: Other)", "Standardized numeric features")
)
print(data_dictionary)
```

## 1.10 Save the cleaned dataset

The cleaned and encoded dataset is written once and serves as the **single
source of truth** for Tasks 3–5 (ANOVA, regression, logistic regression).

```{r}
write.csv(data, "dataset/ds_salaries_clean.csv", row.names = FALSE)
```
