---
title: "Activity 1 — Task 2: Exploratory Data Analysis"
author: "Group 08"
date: "2026-08-16"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = TRUE)
```

This task analyzes the cleaned dataset `ds_salaries_clean.csv` produced in
Task 1 (565 rows, no missing values, no duplicates). The response variable
throughout is `salary_in_usd`. Because the raw salary is strongly right-skewed
(mean 112,298 vs. median 101,570, max 600,000 USD), every comparison below is
shown **twice, side by side**: raw salary in USD (directly interpretable, but
the long tail compresses the boxes) and `log_salary = ln(salary)` (symmetric,
so group differences are visually comparable).

```{r}
library(tidyverse)
data <- read.csv("dataset/ds_salaries_clean.csv")
```

Helper that renders the raw-salary and ln-salary boxplots of one factor
next to each other. Factor levels are relabeled from short codes to full,
readable names so the plots are self-explanatory (see the abbreviation table
in §2.1):

```{r}
label_map <- list(
  experience_level = c(EN = "Entry-level", MI = "Mid-level",
                       SE = "Senior-level", EX = "Executive"),
  employment_type  = c(FT = "Full-time", PT = "Part-time",
                       CT = "Contract", FL = "Freelance"),
  company_size     = c(S = "Small", M = "Medium", L = "Large"),
  location_group   = c(US = "United States", GB = "United Kingdom",
                       CA = "Canada", DE = "Germany", IN = "India",
                       FR = "France", ES = "Spain", Other = "Other"),
  job_category     = c(Data_Analyst = "Data Analyst",
                       Data_Scientist = "Data Scientist",
                       Data_Engineer = "Data Engineer",
                       ML_AI_Engineer = "ML/AI Engineer",
                       Management_Lead = "Management/Lead",
                       Research = "Research", Other = "Other")
)

salary_compare <- function(x_var, x_label, flip = FALSE) {
  df <- data
  if (!is.factor(df[[x_var]])) {
    df[[x_var]] <- factor(df[[x_var]])
  }
  if (x_var %in% names(label_map)) {
    df[[x_var]] <- factor(df[[x_var]],
                          levels = names(label_map[[x_var]]),
                          labels = unname(label_map[[x_var]]))
  }
  p_raw <- ggplot(df, aes(x = .data[[x_var]], y = salary_in_usd)) +
    geom_boxplot(fill = "skyblue", outlier.color = "red") +
    scale_y_continuous(labels = scales::dollar) +
    theme_minimal() +
    labs(title = "Raw Salary (USD)", x = x_label, y = "Salary in USD")

  p_log <- ggplot(df, aes(x = .data[[x_var]], y = log_salary)) +
    geom_boxplot(fill = "skyblue", outlier.color = "red") +
    theme_minimal() +
    labs(title = "ln(Salary)", x = x_label, y = "ln(Salary in USD)")

  if (flip) {
    p_raw <- p_raw + coord_flip()
    p_log <- p_log + coord_flip()
  }
  cowplot::plot_grid(p_raw, p_log, ncol = 2)
}
```

# 2 Exploratory Data Analysis

## 2.1 Abbreviations

The raw data uses short codes for its categorical variables. The following
table decodes every code that appears in the plots below.

| Abbreviation | Meaning |
|---|---|
| FT / PT / CT / FL | Full-time / Part-time / Contract / Freelance employment |
| EN / MI / SE / EX | Entry-level / Mid-level / Senior-level (Expert) / Executive-level (Director) experience |
| S / M / L | Small (< 50 employees) / Medium (50–250) / Large (> 250) company size |
| US / GB / CA / DE / IN / FR / ES | United States / United Kingdom / Canada / Germany / India / France / Spain |
| 0 / 50 / 100 (remote ratio) | No remote work (< 20%) / Partially remote / Fully remote |
| Other (location) | Any country outside the top 7 above |

## 2.2 Descriptive statistics

```{r}
summary(data$salary_in_usd)
```

```{r}
data %>%
  group_by(job_category) %>%
  summarise(n = n(),
            mean_salary = mean(salary_in_usd),
            median_salary = median(salary_in_usd),
            sd_salary = sd(salary_in_usd),
            .groups = "drop") %>%
  arrange(desc(mean_salary))
```

## 2.3 Distribution of salary (raw vs. log)

```{r}
p_raw <- ggplot(data, aes(x = salary_in_usd)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 30) +
  scale_x_continuous(labels = scales::dollar) +
  theme_minimal() +
  labs(title = "Distribution of Salary (USD)", x = "Salary in USD", y = "Count")

p_log <- ggplot(data, aes(x = log_salary)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of ln(Salary)", x = "ln(Salary in USD)", y = "Count")

cowplot::plot_grid(p_raw, p_log, ncol = 2)
```

The raw salary is strongly right-skewed (long tail of high earners up to
600,000 USD), while the log-transformed salary is close to symmetric — which is
why the modeling tasks use `log_salary` as the response.

## 2.4 Salary by job category

```{r}
salary_compare("job_category", "Job Category", flip = TRUE)
```

## 2.5 Salary by experience level

```{r}
salary_compare("experience_level", "Experience Level")
```

## 2.6 Salary by company size

```{r}
salary_compare("company_size", "Company Size")
```

## 2.7 Salary by employment type

```{r}
salary_compare("employment_type", "Employment Type")
```

## 2.8 Salary by remote work ratio (discrete: 0/50/100)

```{r}
salary_compare("remote_ratio", "Remote Ratio (%)")
```

## 2.9 Salary by company location group

```{r}
salary_compare("location_group", "Company Location", flip = TRUE)
```

## 2.10 Salary trend over time (work-year)

```{r}
salary_compare("work_year", "Work Year")
```

## 2.11 Number of jobs per category

```{r}
df_bar <- data %>%
  mutate(job_category = factor(job_category,
                               levels = names(label_map$job_category),
                               labels = unname(label_map$job_category)))
ggplot(df_bar, aes(x = job_category)) +
  geom_bar(fill = "skyblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Number of Jobs per Category", x =
         "", y = "Count")
```