---
title: "Activity 1 — Task 2: Exploratory Data Analysis"
author: "Group 08"
date: "2026-08-16"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = TRUE)
```

This task analyzes the cleaned dataset `ds_salaries_clean.csv` produced in
Task 1 (565 rows, no missing values, no duplicates). The response variable
throughout is `salary_in_usd`; `log_salary` is used for plots because the raw
salary is strongly right-skewed.

```{r}
library(tidyverse)
data <- read.csv("dataset/ds_salaries_clean.csv")
```

# 2 Exploratory Data Analysis

## 2.1 Descriptive statistics

```{r}
summary(data$salary_in_usd)
```

```{r}
data %>%
  group_by(job_category) %>%
  summarise(n = n(),
            mean_salary = mean(salary_in_usd),
            median_salary = median(salary_in_usd),
            sd_salary = sd(salary_in_usd),
            .groups = "drop") %>%
  arrange(desc(mean_salary))
```

## 2.2 Distribution of salary (raw vs. log)

```{r}
ggplot(data, aes(x = salary_in_usd)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 30) +
  scale_x_continuous(labels = scales::dollar) +
  theme_minimal() +
  labs(title = "Distribution of Salary (USD)", x =
         "Salary in USD", y = "Count")

ggplot(data, aes(x = log_salary)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of ln(Salary)", x =
         "ln(Salary in USD)", y = "Count")
```

The raw salary is strongly right-skewed (long tail of high earners up to
600,000 USD), while the log-transformed salary is close to symmetric — which is
why later modeling tasks use `log_salary`.

## 2.3 Salary by job category

```{r}
ggplot(data, aes(x = job_category, y = log_salary)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "ln(Salary) by Job Category", x = "", y = "ln(Salary in USD)")
```

## 2.4 Salary by experience level

```{r}
ggplot(data, aes(x = experience_level, y = log_salary)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  theme_minimal() +
  labs(title = "ln(Salary) by Experience Level", x =
         "Experience Level", y = "ln(Salary in USD)")
```

## 2.5 Salary by company size

```{r}
ggplot(data, aes(x = company_size, y = log_salary)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  theme_minimal() +
  labs(title = "ln(Salary) by Company Size", x =
         "Company Size", y = "ln(Salary in USD)")
```

## 2.6 Salary by employment type

```{r}
ggplot(data, aes(x = employment_type, y = log_salary)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  theme_minimal() +
  labs(title = "ln(Salary) by Employment Type", x =
          "Employment Type", y = "ln(Salary in USD)")
```

## 2.7 Salary by remote work ratio (discrete: 0/50/100)

```{r}
ggplot(data, aes(x = factor(remote_ratio), y = log_salary)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  theme_minimal() +
  labs(title = "ln(Salary) by Remote Work Ratio", x =
         "Remote Ratio (%)", y = "ln(Salary in USD)")
```

## 2.8 Salary by company location group

```{r}
ggplot(data, aes(x = location_group, y = log_salary)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "ln(Salary) by Company Location", x =
          "", y = "ln(Salary in USD)")
```

## 2.9 Salary trend over time (work-year)

```{r}
ggplot(data, aes(x = factor(work_year), y = log_salary)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  theme_minimal() +
  labs(title = "ln(Salary) by Work Year", x =
         "Work Year", y = "ln(Salary in USD)")
```

## 2.10 Number of jobs per category

```{r}
ggplot(data, aes(x = job_category)) +
  geom_bar(fill = "skyblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Number of Jobs per Category", x =
         "", y = "Count")
```
