---
title: "Activity 1 — Task 4: Continuous Salary Regression Modeling"
author: "Group 08"
date: "2026-08-24"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

This section addresses the first research question: *how accurately can we
predict the salary of a Data Science professional worldwide, and which factors
drive income levels?* We model `salary_in_usd` through its natural logarithm
`log_salary`, because the exploratory analysis showed a strongly right-skewed
distribution with a handful of legitimate executive-level salaries; working on
the log scale accommodates that skew without discarding real observations, and
it gives coefficients a convenient percentage interpretation.

Seven model specifications are constructed and their assumptions tested: a
**multiple linear regression**, first with all available predictors and then
reduced by stepwise selection; a **polynomial regression** that lets the
numeric predictors bend away from a straight line; **Ridge**, **LASSO**, and
their **elastic-net** mixture with a cross-validated mixing weight, which
stabilize estimates when many dummy variables share the workload; and a final
linear model fitted on the Box–Cox-optimal power transform
$\text{salary}^{\lambda}$. All fitting, tuning, and assumption testing use
the training data only; the held-out records are kept untouched here and are
formally scored and compared in Task 6.

```{r}
library(tidyverse)
library(glmnet)
library(car)
library(lmtest)
library(MASS)
library(tseries)   # jarque.bera.test
library(sandwich)  # HC-robust standard errors
library(broom)     # tidy coefficient tables

data <- read.csv("dataset/ds_salaries_clean.csv")

set.seed(2026)     # fixed seed: every randomized step below is reproducible
```

# 4 Continuous Salary Regression Modeling

## 4.1 Training and testing partitions

Model quality is judged on observations not seen during fitting, so we split
the 565 cleaned records into training (80%) and test (20%) sets. The first
research question asks how accurately salaries can be predicted **globally**
and which factors drive income across the whole sample, so every observation
era has an equal chance of being held out: a single random 80 : 20 split of
the 565 records assigns 452 rows to training and 113 to test, preserving the
global mix of years, roles, and locations in both partitions. No temporal or
year-aware constraint is imposed, because the goal is a model for the
population of data-science salaries as a whole, not a forecast of any single
future year.

The draw is checked so that rare categories (contract and freelance
employment have 5 and 4 records dataset-wide) stay represented in training;
the fixed seed makes the split fully reproducible.

```{r split-4-1}
used_cols <- c("log_salary", "salary_in_usd", "experience_level", "employment_type",
               "company_size", "job_category", "location_group", "remote_ratio",
               "is_local", "work_year",
               "exp_level_ord_scaled", "company_size_ord_scaled", "remote_ratio_scaled")
stopifnot(nrow(data) == 565, sum(is.na(data[, intersect(used_cols, names(data))])) == 0)

# Factor baselines chosen for interpretability: coefficients then read as
# contrasts against the common category (FT employment, Data Analyst,
# rest-of-world location).
data$experience_level <- factor(data$experience_level, levels = c("EN", "MI", "SE", "EX"))
data$company_size     <- factor(data$company_size,     levels = c("S", "M", "L"))
data$employment_type  <- relevel(factor(data$employment_type), ref = "FT")
data$job_category     <- relevel(factor(data$job_category),   ref = "Data_Analyst")
data$location_group   <- relevel(factor(data$location_group), ref = "Other")

data <- data %>% mutate(row_id = row_number())

# Plain random 80:20 split across the whole dataset (global model, seed 2026).
# Rare categories must stay represented in training, so the draw is re-tried
# until every test-category level also appears among the training rows.
check_vars <- c("employment_type", "job_category", "location_group",
                "experience_level", "company_size")

repeat {
  idx <- sample(nrow(data), floor(0.8 * nrow(data)))
  tr  <- data[idx, ]
  te  <- data[-idx, ]
  ok  <- all(vapply(check_vars,
                    function(v) all(unique(te[[v]]) %in% unique(tr[[v]])), logical(1)))
  if (ok) break
}
train <- tr
test  <- te

stopifnot(nrow(train) == 452, nrow(test) == 113,
          all(!test$row_id %in% train$row_id))

write.csv(train, "dataset/train_data.csv", row.names = FALSE)
write.csv(test, "dataset/test_data.csv", row.names = FALSE)
```

Both partitions preserve a broad global mix of years, experience levels, job
families, and employment types. The rare `CT` category appears only in
training under this seed, while every category observed in the test set is
represented in training, as the following composition tables show:

```{r}
table(train$work_year);        table(test$work_year)
table(train$experience_level); table(test$experience_level)
table(train$job_category);     table(test$job_category)
table(train$employment_type);  table(test$employment_type)
```

## 4.2 Multiple Linear Regression

We build the regression the way an analyst should: start from the simplest
defensible choice — an ordinary linear model on the **raw dollar salary** —
confront it with the classical assumptions, let a Box–Cox analysis select a
better response scale, and then verify that the corrected model genuinely
behaves better. Only after the response scale is settled do we refine the
model and interpret its coefficients.

### A reference model on raw dollars

All working models in this section share the same right-hand side — experience
level, job category, employment type, company size, grouped employer location,
remote-work ratio, survey year, and `is_local` (1 when the employer country
matches the employee's country of residence, 0 otherwise). The reference version
below regresses the raw dollar amount on the demographic design *without*
`is_local`, purely to choose the Box–Cox response scale; every working model that
follows (log, power, and penalized) carries the full set including `is_local`, so
the residence-match signal competes fairly with the other predictors.

```{r raw-reference}
ols_raw <- lm(salary_in_usd ~ experience_level + job_category + employment_type +
                company_size + location_group + remote_ratio + work_year,
              data = train)
summary(ols_raw)
```

On its own scale the model is respectable at first glance (adjusted R² =
0.462), but fit quality is not what concerns us — the assumptions are.

### Diagnostic checking of the reference model

The four base residual plots, formal normality tests, a Breusch–Pagan test
for homoskedasticity, and Cook's distance (influence of single observations)
are applied so that every assumption is
*tested*, never assumed.

```{r raw-diagnostics}
diag_row <- function(label, mod) {
  sw <- shapiro.test(resid(mod))
  jb <- tseries::jarque.bera.test(resid(mod))
  bp <- lmtest::bptest(mod)
  tibble(response       = label,
         Shapiro_p      = signif(sw$p.value, 3),
         Jarque_Bera_p  = signif(jb$p.value, 3),
         BP_stat        = round(unname(bp$statistic), 1),
         BP_p           = signif(bp$p.value, 3),
         Max_Cooks_D    = round(max(cooks.distance(mod)), 3))
}
diag_raw <- diag_row("salary_in_usd", ols_raw)
diag_raw

par(mfrow = c(2, 2)); plot(ols_raw); par(mfrow = c(1, 1))
```

Every check fails. Residual normality is rejected at astronomical
significance (*p* ≈ 1.3·10⁻²¹), errors are heteroskedastic (BP = 41.6,
*p* = 0.010), and the largest Cook's distance reaches 0.18 — a handful of
executive-level records visibly dominate the fitted line. An untransformed
linear model is therefore not a trustworthy basis for inference or
prediction.

### Choosing the response scale: Box–Cox analysis

The Box–Cox family asks which power $\text{salary}^{\lambda}$ makes the
linear model most consistent with its assumptions:

```{r boxcox-scale}
bc <- MASS::boxcox(ols_raw, lambda = seq(-2, 2, length = 100), plotit = TRUE)
lambda_hat <- bc$x[which.max(bc$y)]
## Profile-likelihood 95% confidence interval for lambda
cut_off  <- max(bc$y) - qchisq(0.95, 1) / 2
lambda_ci <- range(bc$x[bc$y >= cut_off])
cat("Box-Cox optimal lambda:", round(lambda_hat, 3), "\n")
cat("95% CI: [", round(lambda_ci[1], 3), ",", round(lambda_ci[2], 3), "]\n")
```

The likelihood peaks at $\hat\lambda \approx 0.26$, with a 95%
profile-likelihood interval $[0.22,\ 0.34]$ that excludes **both** special
cases: $\lambda = 1$ (raw dollars, rejected above) and $\lambda = 0$ (the
log). Formally, the data prefer an intermediate power. We adopt **ln(salary)**
as the working response for its interpretable percentage coefficients, and
give $\text{salary}^{0.26}$ the identical battery of assumption tests below;
whether the intermediate power rivals ln(salary) out of sample is deferred to
Task 6, which owns the held-out comparison.

### The regression on ln(salary)

$$\ln(\text{salary}_i) = \beta_0 + \beta_1\,\text{experience}_i +
\beta_2\,\text{role}_i + \beta_3\,\text{contract type}_i +
\beta_4\,\text{company size}_i + \beta_5\,\text{employer location}_i +
\beta_6\,\text{remote ratio}_i + \beta_7\,\text{year}_i +
\beta_8\,\text{is\_local}_i + \varepsilon_i$$

Location enters through the grouped employer-country variable plus a single
`is_local` flag that records whether the employer country matches the
employee's country of residence. The full employee-residence dummies are
deliberately left out: for the largely non-remote workforce the two countries
coincide, and carrying both sets of dummies would introduce near-duplicate
predictors whose inflated variances distort every coefficient's standard
error. The one-column indicator captures the essence of the residence signal
at negligible collinearity cost.

```{r ols-full}
ols_full <- lm(log_salary ~ experience_level + job_category + employment_type +
                 company_size + location_group + remote_ratio + work_year + is_local,
               data = train)
summary(ols_full)
```

The transformed model explains about **64% of the variation in log-salary**
(adjusted R² = 0.638) and is highly significant overall (*F* = 34.1 on 24 and
427 df, *p* < 10⁻¹⁶) — a clear gain over the raw-scale reference, and a step
up from the same design without `is_local` (adjusted R² = 0.619).

### Re-checking the assumptions after the transformation

```{r ln-diagnostics}
## The Box-Cox-optimal power scale receives the identical battery of tests,
## so the transformation choice is judged on evidence rather than habit.
z_train  <- train$salary_in_usd^lambda_hat
ols_power <- lm(z_train ~ experience_level + job_category + employment_type +
                  company_size + location_group + remote_ratio + work_year + is_local,
                data = train)
diag_power <- diag_row(paste0("salary^", round(lambda_hat, 2)), ols_power)
diag_log <- diag_row("ln(salary)", ols_full)
bind_rows(diag_raw, diag_power, diag_log)

par(mfrow = c(2, 2)); plot(ols_full); par(mfrow = c(1, 1))

## Generalized VIFs: a property of the predictor design matrix, hence
## identical under any response scale — one table covers both models.
vif_raw <- car::vif(ols_full)
if (is.matrix(vif_raw)) {
  vif_tab <- tibble(
    term     = rownames(vif_raw),
    GVIF     = vif_raw[, "GVIF"],
    Df       = vif_raw[, "Df"],
    gvif_adj = vif_raw[, "GVIF^(1/(2*Df))"]
  )
} else {
  vif_tab <- tibble(term = names(vif_raw), GVIF = as.numeric(vif_raw),
                    Df = 1, gvif_adj = sqrt(as.numeric(vif_raw)))
}
vif_tab
```

Side by side, the log transformation earns its place: the Shapiro–Wilk
p-value rises from 1.3·10⁻²¹ (raw dollars) to 3.0·10⁻¹³ — a roughly
two-hundred-and-thirty-million-fold improvement in consistency with
normality — and the maximum Cook's distance falls by a factor of about three
(0.180 → 0.060), so no single record any longer dictates the fit. The
Box–Cox-optimal power behaves comparably (normality *p* = 1.9·10⁻⁸, maximum
Cook's distance 0.072) without clearly outdoing the log on any criterion,
and — like the log scale — its Breusch–Pagan statistic remains significant
(42.2, *p* = 0.012; log: 54.7, *p* = 3.4·10⁻⁴; raw: 41.6, *p* = 0.010). Two
conclusions follow. First, the transformations repair the worst failures of
the raw-dollar model, but residual non-normality and heteroskedasticity
persist — the rejections trace to the thin cluster of very low salaries
documented in the EDA — so robust inference is required regardless of scale.
Second, because neither transformation attains clean diagnostics on every
criterion, the working response is chosen by interpretability — ln(salary)
and its percentage coefficients, used throughout this section — while Task 6
holds the empirical verdict on which scale predicts best. On the
collinearity side, all df-adjusted GVIFs lie between 1.03 and 1.26 — far
under any common threshold — confirming that the single location route plus
the sparse `is_local` flag avoided redundancy.

### Refining the model by AIC

The full model contains two linear trend terms (`remote_ratio`, `work_year`)
whose usefulness a parsimonious model should justify. Stepwise selection in
both directions, guided by Akaike's information criterion, removes exactly
the survey-year term while keeping `remote_ratio`, all five factor blocks,
and the `is_local` flag: the AIC improves from 654.5 to 652.5 at unchanged
adjusted R² (0.638), i.e., the simpler model predicts equally well without
the year trend.

```{r ols-stepwise}
ols_step <- step(ols_full, direction = "both", trace = FALSE)
summary(ols_step)

cat("\nTerms dropped (full model -> stepwise model):",
    setdiff(attr(terms(ols_full), "term.labels"),
            attr(terms(ols_step), "term.labels")), "\n")

tibble(
  model   = c("Full OLS", "Stepwise-reduced OLS"),
  AIC     = c(AIC(ols_full), AIC(ols_step)),
  adj_R2  = c(summary(ols_full)$adj.r.squared, summary(ols_step)$adj.r.squared),
  n_terms = c(length(attr(terms(ols_full), "term.labels")),
              length(attr(terms(ols_step), "term.labels")))
)
```

### Robust inference under unequal error variances

Unequal error variances invalidate ordinary standard errors, so the
coefficient table below is accompanied by HC1-robust
(heteroskedasticity-consistent) ones:

```{r robust-inference}
robust_se <- lmtest::coeftest(ols_step, vcov = sandwich::vcovHC(ols_step, type = "HC1"))
cat("HC1-robust inference for the stepwise-selected ln(salary) model:\n")
print(robust_se)
```

The robust standard errors leave the significance picture essentially
unchanged: every driver flagged as significant below remains significant
under the robust (HC1) errors, so the substantive conclusions are not an
artifact of unequal error variances.

### What drives salaries?

Because the response is a logarithm, each coefficient translates into an
approximate percentage change in salary through $100\,(e^{\beta}-1)$:

```{r salary-drivers}
coef_pct <- broom::tidy(ols_step) %>%
  mutate(pct_effect = 100 * (exp(estimate) - 1)) %>%
  dplyr::select(term, estimate, p.value, pct_effect)
print(coef_pct, n = 30)
```

Four patterns stand out.

**Experience dominates.** Relative to an entry-level professional, mid-level
colleagues earn about 24% more, seniors about 65% more, and executives about
157% more — a steep, strictly increasing gradient that makes seniority the
single strongest career-side lever in the data.

**Role matters within the field.** Against the Data-Analyst baseline, moving
into engineering (+33%) or science (+39%) pays moderately; specializing in
machine learning/AI (+50%), leading teams (+59%), or research (+70%) pays
substantially more. The heterogeneous "Other" category is statistically
indistinguishable from analysts, which supports keeping it separate rather
than forcing it into another family.

**Contract structure shapes pay.** Part-time (−67%) and freelance (−66%)
arrangements earn far less than full-time contracts, consistent with
partial-year or project-based engagement; contract work shows a positive but
unreliable premium (+31%), unsurprising given only five such records.

**Residence-match carries its own premium.** Being employed where you live is
worth roughly **51% more** than an otherwise identical posting at an overseas
employer (*p* = 2·10⁻⁶ ordinary, *p* = 0.0020 robust) — the regression's
confirmation that the cross-border gap Task 6 finds on the same split is a
real structural feature, not an artifact of which countries appear.

**Geography sets the level.** Holding profession and residence-match constant,
a US-based employer pays roughly 155% above the rest-of-world baseline — by
far the largest single effect in the model — with Canada (+88%), Great
Britain (+69%), Germany (+68%) and France (+37%) forming a clear second tier,
while India-based employers pay about 52% below it. Spain is indistinguishable
from the baseline. Company size matters less: large firms pay about 35% more
than small ones (*p* < 0.001); medium firms sit at +17%, significant only on
the ordinary errors (*p* = 0.030 ordinary, *p* = 0.083 robust).

## 4.3 Polynomial Regression

Stepwise selection dropped the survey-year trend but kept `remote_ratio`;
that raises a subtler question: is the remaining trend relationship curved?
Polynomial regression answers it by allowing `remote_ratio` and `work_year`
each to bend instead of following a straight line. Both variables take exactly three distinct values
(0/50/100 and 2020/2021/2022), so degree 3 is not identifiable; degree 2 is
the ceiling and effectively lets every level carry its own effect. We compare
the linear and quadratic specifications with the nested *F*-test.

```{r polynomial}
base_terms <- setdiff(attr(terms(ols_step), "term.labels"), c("remote_ratio", "work_year"))
base_formula_str <- paste("log_salary ~", paste(base_terms, collapse = " + "))

fit_poly <- function(degree) {
  f <- as.formula(paste0(base_formula_str,
                          " + poly(remote_ratio, ", degree, ")",
                          " + poly(work_year, ", degree, ")"))
  lm(f, data = train)
}

poly_linear    <- fit_poly(1)   # straight-line benchmark
poly_quadratic <- fit_poly(2)   # quadratic alternative

tibble(
  degree = c(1, 2),
  AIC    = c(AIC(poly_linear), AIC(poly_quadratic)),
  adj_R2 = c(summary(poly_linear)$adj.r.squared, summary(poly_quadratic)$adj.r.squared)
)

anova_poly <- anova(poly_linear, poly_quadratic)
anova_poly

poly_model <- if (anova_poly$`Pr(>F)`[2] < 0.05) poly_quadratic else poly_linear

reset_result <- lmtest::resettest(poly_model)
cat("\nRESET test on selected model: F =", round(reset_result$statistic, 3),
    " p =", signif(reset_result$p.value, 4), "\n")
```

Out of sample the choice matters, but on the training data the quadratic
specification is not supported: the *F*-test fails to reject the purely
linear form (*F* = 2.48, *p* = 0.085), adjusted R² barely edges up from 0.638
to 0.640, and the RESET test finds no remaining functional-form
misspecification on the selected degree-1 model (*F* = 1.99, *p* = 0.138). In
substantive terms there is no evidence that fully-remote and fully on-site
arrangements sit off the straight line connecting the extremes — the 50%
hybrid point behaves broadly like the average of the two ends — so the
parsimonious linear version is carried forward as the polynomial candidate
for Task 6.

## 4.4 Ridge and LASSO Regression

Ordinary least squares treats all 21 columns as free parameters.
Shrinkage methods trade a little bias for variance reduction and, in the
LASSO's case, automatic variable selection — a useful cross-check on the
stepwise route. They require standardized inputs: all four numeric predictors
are z-scored with training means and standard deviations, applied unchanged
to the test records.

```{r ridge-lasso-matrices}
## Training-only z-scaling of every numeric input; the identical transform
## is applied to test.
tr_scale <- function(col) {
  m <- mean(train[[col]]); s <- sd(train[[col]])
  list(train = (train[[col]] - m) / s, test = (test[[col]] - m) / s)
}
for (col in c("exp_level_ord", "company_size_ord", "remote_ratio", "work_year")) {
  sc <- tr_scale(col)
  train[[paste0(col, "_ts")]] <- sc$train
  test[[paste0(col, "_ts")]]  <- sc$test
}

build_matrix <- function(df) {
  model.matrix(
    ~ exp_level_ord_ts + company_size_ord_ts + remote_ratio_ts +
      work_year_ts + job_category + employment_type + location_group + is_local,
    data = df
  )[, -1]   # glmnet supplies its own intercept
}

x_train <- build_matrix(train)
x_test  <- build_matrix(test)
y_train <- train$log_salary
y_test  <- test$log_salary
dim(x_train); dim(x_test)
```

The penalty strength λ is tuned by ten-fold cross-validation within the
training set; prediction uses `lambda.min`, with the sparser `lambda.1se`
reported for reference.

```{r ridge-lasso-cv}
set.seed(2026)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)
set.seed(2026)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

cat("Ridge: lambda.min =", cv_ridge$lambda.min, " lambda.1se =", cv_ridge$lambda.1se, "\n")
cat("LASSO: lambda.min =", cv_lasso$lambda.min, " lambda.1se =", cv_lasso$lambda.1se, "\n")

plot(cv_ridge); title("Ridge cross-validation curve", line = 2.5)
plot(cv_lasso); title("LASSO cross-validation curve", line = 2.5)

## Robustness of the endpoint choice: a sweep over the full elastic-net
## mixing range finds the CV surface flat, so neither pure penalty is a
## lucky guess.
enet_sweep <- vapply(seq(0, 1, by = 0.1), function(a) {
  set.seed(2026)
  sqrt(min(cv.glmnet(x_train, y_train, alpha = a, nfolds = 10)$cvm))
}, numeric(1))
cat("CV RMSE across alpha = 0,0.1,...,1:",
    sprintf("%.4f-%.4f (range %.4f)", min(enet_sweep), max(enet_sweep),
            diff(range(enet_sweep))), "\n")

plot(seq(0, 1, by = 0.1), enet_sweep, type = "b", pch = 19,
     xlab = expression(alpha * "  (0 = pure Ridge, 1 = pure LASSO)"),
     ylab = "Cross-validated RMSE (log points)",
     main = "Elastic-net mixing sweep")

## Tuned elastic net: alpha picked by the same 10-fold CV, kept alongside its
## two pure-parent special cases for Task 6 to score.
alpha_best <- seq(0, 1, by = 0.1)[which.min(enet_sweep)]
set.seed(2026)
cv_elastic <- cv.glmnet(x_train, y_train, alpha = alpha_best, nfolds = 10)
cat("Selected elastic-net alpha:", alpha_best,
    "| predictor columns kept at lambda.min:",
    sum(as.numeric(coef(cv_elastic, s = "lambda.min"))[-1] != 0), "/", ncol(x_train), "\n")
```

Sweeping the mixing parameter across its full range ($\alpha$ = 0, 0.1, …, 1)
moves the cross-validated error by about 0.16%, and nearly every column
survives at $\lambda_{\min}$ (19–21 of the 21 columns), with the flat
`work_year_ts` and `location_groupES` effects the only ones ever shrunk all
the way to zero. With the surface this flat there is no meaningful interior
optimum; the CV-selected mixture
($\alpha$ = `r round(alpha_best, 1)`) is carried forward for Task 6 to score
alongside the other specifications.

Which variables survive the LASSO penalty, and do they agree with the stepwise
choice? Since LASSO names individual dummy columns while stepwise names whole
variables, survivors are first mapped back to their parent variables before
comparing.

```{r lasso-variable-selection}
lasso_coefs_min <- coef(cv_lasso, s = "lambda.min")
lasso_survivors <- setdiff(rownames(lasso_coefs_min)[which(lasso_coefs_min[, 1] != 0)],
                           "(Intercept)")
cat("LASSO retains", length(lasso_survivors), "of", nrow(lasso_coefs_min) - 1,
    "columns at lambda.min\n")

parent_terms <- c("exp_level_ord_ts", "company_size_ord_ts",
                  "remote_ratio_ts", "work_year_ts",
                  "job_category", "employment_type", "location_group",
                  "is_local")
map_to_parent <- function(dummy_names) {
  vapply(dummy_names, function(nm) {
    hit <- parent_terms[startsWith(nm, parent_terms)]
    if (length(hit) == 0) nm else hit[which.max(nchar(hit))]
  }, character(1), USE.NAMES = FALSE)
}
canonical <- c(exp_level_ord_ts         = "experience_level",
               company_size_ord_ts      = "company_size",
               remote_ratio_ts          = "remote_ratio",
               work_year_ts             = "work_year")
lasso_parents <- unique(map_to_parent(lasso_survivors))
lasso_parents <- unique(ifelse(lasso_parents %in% names(canonical),
                               canonical[lasso_parents], lasso_parents))
stepwise_terms <- attr(terms(ols_step), "term.labels")

cat("Selected by BOTH:", intersect(stepwise_terms, lasso_parents), "\n")
cat("Stepwise only:",     setdiff(stepwise_terms, lasso_parents), "\n")
cat("LASSO only:        ", setdiff(lasso_parents, stepwise_terms), "\n")
```

The two selection philosophies now converge exactly: both retain the same
seven blocks — experience level, job family, contract type, company size,
location, remote ratio, and the `is_local` flag — and both discard the
survey-year term (AIC by dropping `work_year`, LASSO by shrinking
`work_year_ts` exactly to zero alongside the flat `location_groupES` dummy).
The agreement is no longer cosmetic: stepwise and LASSO nominate the
identical variable set.

A final sensitivity check replaces the ordinal encodings of experience and
company size with full dummies inside the LASSO, the two variants being
compared by ten-fold cross-validation on the training data. Accuracy hardly
moves (CV RMSE 0.5055 vs 0.5077): the ordinal simplification costs nothing
predictive.

```{r sensitivity-encoding}
build_matrix_factor <- function(df) {
  model.matrix(
    ~ experience_level + company_size + remote_ratio_ts + work_year_ts +
      job_category + employment_type + location_group + is_local,
    data = df
  )[, -1]
}
x_train_f <- build_matrix_factor(train)

set.seed(2026)
cv_lasso_factor <- cv.glmnet(x_train_f, y_train, alpha = 1, nfolds = 10)

cat("LASSO CV RMSE (log scale, training folds):\n")
cat("  ordinal encoding:", round(sqrt(min(cv_lasso$cvm)), 4),
    "| factor-dummy encoding:", round(sqrt(min(cv_lasso_factor$cvm)), 4), "\n")
```

## 4.5 Summary of training-set fit

This task constructs the candidate models, verifies their assumptions, and
interprets the drivers on the training data; formal scoring of the held-out
records and the final model choice belong to Task 6. To close the modeling
here, the table below summarizes how well each specification fits the 452
training rows. Because every model except the power transform shares
`log_salary` as its response, their R² and adjusted R² are directly
comparable on the log scale. AIC — which rewards fit while penalizing
parameter count — is reported for the three ordinary least-squares fits
where it is defined; it is not defined for the penalty-regularized
Ridge/LASSO/elastic-net fits, whose effective parameter count is not the
number of columns.

```{r train-fit-comparison}
pred_tr <- list(
  `Full OLS - stepwise`         = predict(ols_step),
  `Full OLS`                    = predict(ols_full),
  `Polynomial regression`       = predict(poly_model),
  `Ridge regression`            = as.numeric(predict(cv_ridge, newx = x_train, s = "lambda.min")),
  `LASSO regression`            = as.numeric(predict(cv_lasso, newx = x_train, s = "lambda.min")),
  `Elastic net regression`      = as.numeric(predict(cv_elastic, newx = x_train, s = "lambda.min"))
)

fit_row <- function(model, fit, y, n_coef, aic_val) {
  n    <- length(y)
  rss  <- sum((y - fit)^2)
  r2   <- 1 - rss / sum((y - mean(y))^2)
  adj  <- 1 - (1 - r2) * (n - 1) / (n - n_coef - 1)
  tibble(model = model,
         R2     = signif(r2, 4),
         adj_R2 = signif(adj, 4),
         n_par  = n_coef,
         AIC    = aic_val)
}

p_full <- length(coef(ols_full))
p_step <- length(coef(ols_step))
p_poly <- length(coef(poly_model))

train_fit <- bind_rows(
  fit_row("Multiple linear regression - full",      pred_tr[["Full OLS"]],            y_train, p_full, round(AIC(ols_full), 1)),
  fit_row("Multiple linear regression - stepwise",  pred_tr[["Full OLS - stepwise"]],  y_train, p_step, round(AIC(ols_step), 1)),
  fit_row("Polynomial regression",                  pred_tr[["Polynomial regression"]], y_train, p_poly, round(AIC(poly_model), 1)),
  fit_row("Ridge regression",                       pred_tr[["Ridge regression"]],      y_train, ncol(x_train), NA),
  fit_row("LASSO regression",                       pred_tr[["LASSO regression"]],      y_train, ncol(x_train), NA),
  fit_row("Elastic net regression",                 pred_tr[["Elastic net regression"]], y_train, ncol(x_train), NA)
)
train_fit

cat("\nNote: the Box-Cox power-scale model (Section 4.2) is fitted on\n")
cat("a different response scale, so its in-sample R2 is not directly\n")
cat("comparable with the log-salary models above and is omitted here; the\n")
cat("held-out comparison across scales is deferred to Task 6.\n")
```

Three points follow. First, the **stepwise-reduced model matches the full
model on training data** — equal adjusted R² (0.638) at a slightly lower AIC
(652.5 vs 654.5) with one fewer parameter — so the pruning costs no in-sample
fit while removing the marginal survey-year trend, the qualitative conclusion
of Section 4.2. Second, the **regularized fits show essentially identical
training R² to the OLS variants** (all in the range 0.651–0.657), as
expected: shrinkage buys stability rather than in-sample fit, and their
advantage, if any, materializes out of sample — which Task 6 measures.
Third, **adjusted R² falls below R² by only a small margin** (about 0.02),
because the models' predictor count is modest relative to 452 training rows,
so none of the candidates is badly over-parameterized in sample. In short,
on the training data the candidate models are nearly indistinguishable in
fit; their separation must come from held-out performance, which Task 6
evaluates.

```{r}
write.csv(train, "dataset/train_data.csv", row.names = FALSE)
write.csv(test, "dataset/test_data.csv", row.names = FALSE)
```

