---
title: "Activity 1 — Task 4: Continuous Salary Regression Modeling"
author: "Group 08"
date: "2026-08-24"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

This section addresses the first research question: *how accurately can we
predict the salary of a Data Science professional worldwide, and which factors
drive income levels?* We model `salary_in_usd` through its natural logarithm
`log_salary`, because the exploratory analysis showed a strongly right-skewed
distribution with a handful of legitimate executive-level salaries; working on
the log scale accommodates that skew without discarding real observations, and
it gives coefficients a convenient percentage interpretation.

Seven models are fitted and compared: a **multiple linear regression**, first
with all available predictors and then reduced by stepwise selection; a
**polynomial regression** that lets the numeric predictors bend away from a
straight line; **Ridge**, **LASSO**, and their **elastic-net** mixture with a
cross-validated mixing weight, which stabilize estimates when many dummy
variables share the workload; and a final linear model fitted on the
Box–Cox-optimal power transform $\text{salary}^{\lambda}$, which lets the
held-out data judge the response scale chosen in Section 4.2. All tuning uses
training data only; the seven candidates are scored together on the held-out
records in Section 4.5.

```{r}
library(tidyverse)
library(glmnet)
library(car)
library(lmtest)
library(MASS)
library(tseries)   # jarque.bera.test
library(sandwich)  # HC-robust standard errors
library(broom)     # tidy coefficient tables

data <- read.csv("activity1/dataset/ds_salaries_clean.csv")

set.seed(2026)     # fixed seed: every randomized step below is reproducible
```

# 4 Continuous Salary Regression Modeling

## 4.1 Training and testing partitions

Model quality is judged on observations not seen during fitting, so we split
the 565 cleaned records into training (80%) and test (20%) sets. Salaries
evolve over time, so the split leans toward recent data: training keeps all
287 records from 2020–2021 plus 165 randomly sampled 2022 records (452 rows),
while the 113-row test set contains only 2022 records — a *year-aware*
rather than strictly chronological split, since about three of five 2022
salaries also inform the fit. Holding out all of 2022 would have left a 49%
test set, so the sample preserves the 80:20 balance.

The draw is checked so that rare categories (contract and freelance
employment have 5 and 4 records dataset-wide) stay represented in training;
the fixed seed makes the split fully reproducible.

```{r split-4-1}
used_cols <- c("log_salary", "salary_in_usd", "experience_level", "employment_type",
               "company_size", "job_category", "location_group", "remote_ratio",
               "is_local", "work_year",
               "exp_level_ord_scaled", "company_size_ord_scaled", "remote_ratio_scaled")
stopifnot(nrow(data) == 565, sum(is.na(data[, intersect(used_cols, names(data))])) == 0)

# Factor baselines chosen for interpretability: coefficients then read as
# contrasts against the common category (FT employment, Data Analyst,
# rest-of-world location).
data$experience_level <- factor(data$experience_level, levels = c("EN", "MI", "SE", "EX"))
data$company_size     <- factor(data$company_size,     levels = c("S", "M", "L"))
data$employment_type  <- relevel(factor(data$employment_type), ref = "FT")
data$job_category     <- relevel(factor(data$job_category),   ref = "Data_Analyst")
data$location_group   <- relevel(factor(data$location_group), ref = "Other")

data <- data %>% mutate(row_id = row_number())

train_old <- data %>% filter(work_year %in% c(2020, 2021))   # 287 older records
pool_2022 <- data %>% filter(work_year == 2022)              # 278 recent records

check_vars <- c("employment_type", "job_category", "location_group",
                "experience_level", "company_size")

repeat {
  idx  <- sample(nrow(pool_2022), 165)
  tr22 <- pool_2022[idx, ]
  te   <- pool_2022[-idx, ]
  tr   <- bind_rows(train_old, tr22)
  ok <- all(vapply(check_vars,
                   function(v) all(unique(te[[v]]) %in% unique(tr[[v]])), logical(1)))
  if (ok) break
}
train <- tr
test  <- te

stopifnot(nrow(train) == 452, nrow(test) == 113, all(test$work_year == 2022))
```

Both partitions cover the same range of experience levels, job families, and
employment types, as the following composition tables show:

```{r}
table(train$work_year);        table(test$work_year)
table(train$experience_level); table(test$experience_level)
table(train$job_category);     table(test$job_category)
table(train$employment_type);  table(test$employment_type)
```

## 4.2 Multiple Linear Regression

We build the regression the way an analyst should: start from the simplest
defensible choice — an ordinary linear model on the **raw dollar salary** —
confront it with the classical assumptions, let a Box–Cox analysis select a
better response scale, and then verify that the corrected model genuinely
behaves better. Only after the response scale is settled do we refine the
model and interpret its coefficients.

### A reference model on raw dollars

All models in this section share the same right-hand side — experience level,
job category, employment type, company size, grouped employer location,
remote-work ratio, and survey year. The reference version regresses the raw
dollar amount on them:

```{r raw-reference}
ols_raw <- lm(salary_in_usd ~ experience_level + job_category + employment_type +
                company_size + location_group + remote_ratio + work_year,
              data = train)
summary(ols_raw)
```

On its own scale the model is respectable at first glance (adjusted R² =
0.476), but fit quality is not what concerns us — the assumptions are.

### Diagnostic checking of the reference model

The four base residual plots, formal normality tests, a Breusch–Pagan test
for homoskedasticity, and Cook's distance (influence of single observations)
are applied so that every assumption is
*tested*, never assumed.

```{r raw-diagnostics}
diag_row <- function(label, mod) {
  sw <- shapiro.test(resid(mod))
  jb <- tseries::jarque.bera.test(resid(mod))
  bp <- lmtest::bptest(mod)
  tibble(response       = label,
         Shapiro_p      = signif(sw$p.value, 3),
         Jarque_Bera_p  = signif(jb$p.value, 3),
         BP_stat        = round(unname(bp$statistic), 1),
         BP_p           = signif(bp$p.value, 3),
         Max_Cooks_D    = round(max(cooks.distance(mod)), 3))
}
diag_raw <- diag_row("salary_in_usd", ols_raw)
diag_raw

par(mfrow = c(2, 2)); plot(ols_raw); par(mfrow = c(1, 1))
```

Every check fails. Residual normality is rejected at astronomical
significance (*p* ≈ 8·10⁻²²), errors are heteroskedastic (BP = 35.8,
*p* = 0.043), and the largest Cook's distance reaches 0.18 — a handful of
executive-level records visibly dominate the fitted line. An untransformed
linear model is therefore not a trustworthy basis for inference or
prediction.

### Choosing the response scale: Box–Cox analysis

The Box–Cox family asks which power $\text{salary}^{\lambda}$ makes the
linear model most consistent with its assumptions:

```{r boxcox-scale}
bc <- MASS::boxcox(ols_raw, plotit = TRUE)
lambda_hat <- bc$x[which.max(bc$y)]
## Profile-likelihood 95% confidence interval for lambda
cut_off  <- max(bc$y) - qchisq(0.95, 1) / 2
lambda_ci <- range(bc$x[bc$y >= cut_off])
cat("Box-Cox optimal lambda:", round(lambda_hat, 3), "\n")
cat("95% CI: [", round(lambda_ci[1], 3), ",", round(lambda_ci[2], 3), "]\n")
```

The likelihood peaks at $\hat\lambda \approx 0.30$, with a 95%
profile-likelihood interval $[0.26,\ 0.38]$ that excludes **both** special
cases: $\lambda = 1$ (raw dollars, rejected above) and $\lambda = 0$ (the
log). Formally, the data prefer an intermediate power. We adopt **ln(salary)**
as the working response for its interpretable percentage coefficients, and
give $\text{salary}^{0.3}$ the identical battery of assumption tests below;
out-of-sample accuracy follows in Section 4.5.

### The regression on ln(salary)

$$\ln(\text{salary}_i) = \beta_0 + \beta_1\,\text{experience}_i +
\beta_2\,\text{role}_i + \beta_3\,\text{contract type}_i +
\beta_4\,\text{company size}_i + \beta_5\,\text{location}_i +
\beta_6\,\text{remote ratio}_i + \beta_7\,\text{year}_i + \varepsilon_i$$

Location enters through the grouped employer-country variable only. The
employee-residence dummies are deliberately left out: for the largely
non-remote workforce the two countries coincide, and carrying both would
introduce near-duplicate predictors whose inflated variances distort every
coefficient's standard error.

```{r ols-full}
ols_full <- lm(log_salary ~ experience_level + job_category + employment_type +
                 company_size + location_group + remote_ratio + work_year,
               data = train)
summary(ols_full)
```

The transformed model explains about **59% of the variation in log-salary**
(adjusted R² = 0.589) and is highly significant overall (*F* = 29.1 on 23 and
428 df, *p* < 10⁻¹⁶) — a clear gain over the raw-scale reference.

### Re-checking the assumptions after the transformation

```{r ln-diagnostics}
## The Box-Cox-optimal power scale receives the identical battery of tests,
## so the transformation choice is judged on evidence rather than habit.
z_train  <- train$salary_in_usd^lambda_hat
ols_power <- lm(z_train ~ experience_level + job_category + employment_type +
                  company_size + location_group + remote_ratio + work_year,
                data = train)
diag_power <- diag_row(paste0("salary^", round(lambda_hat, 2)), ols_power)
diag_log <- diag_row("ln(salary)", ols_full)
bind_rows(diag_raw, diag_power, diag_log)

par(mfrow = c(2, 2)); plot(ols_full); par(mfrow = c(1, 1))

## Generalized VIFs: a property of the predictor design matrix, hence
## identical under any response scale — one table covers both models.
vif_raw <- car::vif(ols_full)
if (is.matrix(vif_raw)) {
  vif_tab <- tibble(
    term     = rownames(vif_raw),
    GVIF     = vif_raw[, "GVIF"],
    Df       = vif_raw[, "Df"],
    gvif_adj = vif_raw[, "GVIF^(1/(2*Df))"]
  )
} else {
  vif_tab <- tibble(term = names(vif_raw), GVIF = as.numeric(vif_raw),
                    Df = 1, gvif_adj = sqrt(as.numeric(vif_raw)))
}
vif_tab
```

Side by side, the log transformation earns its place: the Shapiro–Wilk
p-value improves roughly 250-million-fold (from 8·10⁻²² to 2·10⁻¹³), and the
maximum Cook's distance halves (0.180 → 0.083), so no single record any
longer dictates the fit. The Box–Cox-optimal power does better still on every
criterion the likelihood flagged — normality improves by a further five
orders of magnitude (*p* = 1.3·10⁻⁸), influence drops to 0.075, and in-sample
fit edges ahead (adjusted R² 0.599 vs 0.589) — while its Breusch–Pagan
statistic (39.0) sits well below the log's 58.0, though heteroskedasticity
remains significant on every scale. Two conclusions follow. First, no
transform achieves normality or constant variance — the rejections trace to
the thin cluster of very low salaries documented in the EDA — so robust
inference is required regardless of scale. Second, because the power scale
dominates on diagnostics, keeping ln(salary) needs empirical backing: the
held-out data arbitrate the two transformed models in Section 4.5. On the
collinearity side, all df-adjusted GVIFs lie between 1.03 and 1.21 — far
under any common threshold — confirming that the single location route
avoided redundancy.

### Refining the model by AIC

The full model contains two linear trend terms (`remote_ratio`, `work_year`)
whose usefulness a parsimonious model should justify. Stepwise selection in
both directions, guided by Akaike's information criterion, removes exactly
those two terms while keeping all five factor blocks: the AIC improves from
740 to 738 at unchanged adjusted R², i.e., the simpler model predicts equally
well with two-thirds of the terms.

```{r ols-stepwise}
ols_step <- step(ols_full, direction = "both", trace = FALSE)
summary(ols_step)

cat("\nTerms dropped (full model -> stepwise model):",
    setdiff(attr(terms(ols_full), "term.labels"),
            attr(terms(ols_step), "term.labels")), "\n")

tibble(
  model   = c("Full OLS", "Stepwise-reduced OLS"),
  AIC     = c(AIC(ols_full), AIC(ols_step)),
  adj_R2  = c(summary(ols_full)$adj.r.squared, summary(ols_step)$adj.r.squared),
  n_terms = c(length(attr(terms(ols_full), "term.labels")),
              length(attr(terms(ols_step), "term.labels")))
)
```

### Robust inference under unequal error variances

Unequal error variances invalidate ordinary standard errors, so the
coefficient table below is accompanied by HC1-robust
(heteroskedasticity-consistent) ones:

```{r robust-inference}
robust_se <- lmtest::coeftest(ols_step, vcov = sandwich::vcovHC(ols_step, type = "HC1"))
cat("HC1-robust inference for the stepwise-selected ln(salary) model:\n")
print(robust_se)
```

The robust standard errors leave the significance picture essentially
unchanged; the single exception — medium company size — is flagged below.

### What drives salaries?

Because the response is a logarithm, each coefficient translates into an
approximate percentage change in salary through $100\,(e^{\beta}-1)$:

```{r salary-drivers}
coef_pct <- broom::tidy(ols_step) %>%
  mutate(pct_effect = 100 * (exp(estimate) - 1)) %>%
  dplyr::select(term, estimate, p.value, pct_effect)
print(coef_pct, n = 30)
```

Four patterns stand out.

**Experience dominates.** Relative to an entry-level professional, mid-level
colleagues earn about 26% more, seniors about 75% more, and executives about
168% more — a steep, strictly increasing gradient that makes seniority the
single strongest career-side lever in the data.

**Role matters within the field.** Against the Data-Analyst baseline, moving
into engineering (+25%) or science (+29%) pays moderately; specializing in
machine learning/AI (+47%), leading teams (+48%), or research (+64%) pays
substantially more. The heterogeneous "Other" category is statistically
indistinguishable from analysts, which supports keeping it separate rather
than forcing it into another family.

**Contract structure shapes pay.** Part-time (−48%) and freelance (−64%)
arrangements earn significantly less than full-time contracts, consistent
with partial-year or project-based engagement; contract work shows a positive
but unreliable premium (+42%), unsurprising given only five such records.

**Geography sets the level.** Holding profession constant, a US-based
employer pays roughly 170% above the rest-of-world baseline — by far the
largest single effect in the model — with Canada (+90%), Great Britain
(+85%), Germany (+77%) and France (+42%) forming a clear second tier, while
India-based employers pay about 47% below it. Spain is indistinguishable from
the baseline. Company size matters less: large firms pay about 30% more than
small ones (*p* < 0.001); medium firms sit at +16%, marginal on ordinary
errors (*p* = 0.052) and not significant under robust ones (*p* = 0.125).

## 4.3 Polynomial Regression

Stepwise selection suggested the linear trend terms added little, but that
raises a subtler question: is the relationship curved? Polynomial regression
answers it by allowing `remote_ratio` and `work_year` each to bend instead of
following a straight line. Both variables take exactly three distinct values
(0/50/100 and 2020/2021/2022), so degree 3 is not identifiable; degree 2 is
the ceiling and effectively lets every level carry its own effect. We compare
the linear and quadratic specifications with the nested *F*-test.

```{r polynomial}
base_terms <- setdiff(attr(terms(ols_step), "term.labels"), c("remote_ratio", "work_year"))
base_formula_str <- paste("log_salary ~", paste(base_terms, collapse = " + "))

fit_poly <- function(degree) {
  f <- as.formula(paste0(base_formula_str,
                          " + poly(remote_ratio, ", degree, ")",
                          " + poly(work_year, ", degree, ")"))
  lm(f, data = train)
}

poly_linear    <- fit_poly(1)   # straight-line benchmark
poly_quadratic <- fit_poly(2)   # quadratic alternative

tibble(
  degree = c(1, 2),
  AIC    = c(AIC(poly_linear), AIC(poly_quadratic)),
  adj_R2 = c(summary(poly_linear)$adj.r.squared, summary(poly_quadratic)$adj.r.squared)
)

anova_poly <- anova(poly_linear, poly_quadratic)
anova_poly

poly_model <- if (anova_poly$`Pr(>F)`[2] < 0.05) poly_quadratic else poly_linear

reset_result <- lmtest::resettest(poly_model)
cat("\nRESET test on selected model: F =", round(reset_result$statistic, 3),
    " p =", signif(reset_result$p.value, 4), "\n")
```

The quadratic specification is preferred: the *F*-test rejects the purely
linear form (*F* = 3.60, *p* = 0.028), adjusted R² edges up from 0.589 to
0.594, and the RESET test finds no remaining functional-form misspecification
(*F* = 2.13, *p* = 0.12). In substantive terms, fully-remote and fully
on-site arrangements sit at different pay levels than a straight line through
the middle would suggest — the 50% hybrid point behaves differently from the
average of the extremes.

## 4.4 Ridge and LASSO Regression

Ordinary least squares treats all twenty columns as free parameters.
Shrinkage methods trade a little bias for variance reduction and, in the
LASSO's case, automatic variable selection — a useful cross-check on the
stepwise route. They require standardized inputs: all four numeric predictors
are z-scored with training means and standard deviations, applied unchanged
to the test records.

```{r ridge-lasso-matrices}
## Training-only z-scaling of every numeric input; the identical transform
## is applied to test.
tr_scale <- function(col) {
  m <- mean(train[[col]]); s <- sd(train[[col]])
  list(train = (train[[col]] - m) / s, test = (test[[col]] - m) / s)
}
for (col in c("exp_level_ord", "company_size_ord", "remote_ratio", "work_year")) {
  sc <- tr_scale(col)
  train[[paste0(col, "_ts")]] <- sc$train
  test[[paste0(col, "_ts")]]  <- sc$test
}

build_matrix <- function(df) {
  model.matrix(
    ~ exp_level_ord_ts + company_size_ord_ts + remote_ratio_ts +
      work_year_ts + job_category + employment_type + location_group,
    data = df
  )[, -1]   # glmnet supplies its own intercept
}

x_train <- build_matrix(train)
x_test  <- build_matrix(test)
y_train <- train$log_salary
y_test  <- test$log_salary
dim(x_train); dim(x_test)
```

The penalty strength λ is tuned by ten-fold cross-validation within the
training set; prediction uses `lambda.min`, with the sparser `lambda.1se`
reported for reference.

```{r ridge-lasso-cv}
set.seed(2026)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)
set.seed(2026)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

cat("Ridge: lambda.min =", cv_ridge$lambda.min, " lambda.1se =", cv_ridge$lambda.1se, "\n")
cat("LASSO: lambda.min =", cv_lasso$lambda.min, " lambda.1se =", cv_lasso$lambda.1se, "\n")

plot(cv_ridge); title("Ridge cross-validation curve", line = 2.5)
plot(cv_lasso); title("LASSO cross-validation curve", line = 2.5)

## Robustness of the endpoint choice: a sweep over the full elastic-net
## mixing range finds the CV surface flat, so neither pure penalty is a
## lucky guess.
enet_sweep <- vapply(seq(0, 1, by = 0.1), function(a) {
  set.seed(2026)
  sqrt(min(cv.glmnet(x_train, y_train, alpha = a, nfolds = 10)$cvm))
}, numeric(1))
cat("CV RMSE across alpha = 0,0.1,...,1:",
    sprintf("%.4f-%.4f (range %.4f)", min(enet_sweep), max(enet_sweep),
            diff(range(enet_sweep))), "\n")

plot(seq(0, 1, by = 0.1), enet_sweep, type = "b", pch = 19,
     xlab = expression(alpha * "  (0 = pure Ridge, 1 = pure LASSO)"),
     ylab = "Cross-validated RMSE (log points)",
     main = "Elastic-net mixing sweep")

## Tuned elastic net: alpha picked by the same 10-fold CV, refit for scoring
## in Section 4.5 alongside its two pure-parent special cases.
alpha_best <- seq(0, 1, by = 0.1)[which.min(enet_sweep)]
set.seed(2026)
cv_elastic <- cv.glmnet(x_train, y_train, alpha = alpha_best, nfolds = 10)
cat("Selected elastic-net alpha:", alpha_best,
    "| predictor columns kept at lambda.min:",
    sum(as.numeric(coef(cv_elastic, s = "lambda.min"))[-1] != 0), "/", ncol(x_train), "\n")
```

Sweeping the mixing parameter across its full range ($\alpha$ = 0, 0.1, …, 1)
moves the cross-validated error by less than 0.15%, and every mixture retains
all twenty columns at $\lambda_{\min}$. With the surface this flat there is no
meaningful interior optimum; the CV-selected mixture
($\alpha$ = `r round(alpha_best, 1)`) still joins the Section 4.5 ranking so
that the comparison covers the entire penalty path.

Which variables survive the LASSO penalty, and do they agree with the stepwise
choice? Since LASSO names individual dummy columns while stepwise names whole
variables, survivors are first mapped back to their parent variables before
comparing.

```{r lasso-variable-selection}
lasso_coefs_min <- coef(cv_lasso, s = "lambda.min")
lasso_survivors <- setdiff(rownames(lasso_coefs_min)[which(lasso_coefs_min[, 1] != 0)],
                           "(Intercept)")
cat("LASSO retains", length(lasso_survivors), "of", nrow(lasso_coefs_min) - 1,
    "columns at lambda.min\n")

parent_terms <- c("exp_level_ord_ts", "company_size_ord_ts",
                  "remote_ratio_ts", "work_year_ts",
                  "job_category", "employment_type", "location_group")
map_to_parent <- function(dummy_names) {
  vapply(dummy_names, function(nm) {
    hit <- parent_terms[startsWith(nm, parent_terms)]
    if (length(hit) == 0) nm else hit[which.max(nchar(hit))]
  }, character(1), USE.NAMES = FALSE)
}
canonical <- c(exp_level_ord_ts         = "experience_level",
               company_size_ord_ts      = "company_size",
               remote_ratio_ts          = "remote_ratio",
               work_year_ts             = "work_year")
lasso_parents <- unique(map_to_parent(lasso_survivors))
lasso_parents <- unique(ifelse(lasso_parents %in% names(canonical),
                               canonical[lasso_parents], lasso_parents))
stepwise_terms <- attr(terms(ols_step), "term.labels")

cat("Selected by BOTH:", intersect(stepwise_terms, lasso_parents), "\n")
cat("Stepwise only:",     setdiff(stepwise_terms, lasso_parents), "\n")
cat("LASSO only:        ", setdiff(lasso_parents, stepwise_terms), "\n")
```

The two selection philosophies converge on the same story: both retain the
five substantive blocks — experience level, job family, contract type,
company size, and location. LASSO additionally preserves small linear
`remote_ratio` and `work_year` components that AIC pruning had discarded;
their coefficients are negligible in size, so the disagreement is cosmetic
rather than substantive.

A final sensitivity check replaces the ordinal encodings of experience and
company size with full dummies inside the LASSO, the two variants being
compared by ten-fold cross-validation on the training data. Accuracy hardly
moves (CV RMSE 0.5491 vs 0.5514): the ordinal simplification costs nothing
predictive.

```{r sensitivity-encoding}
build_matrix_factor <- function(df) {
  model.matrix(
    ~ experience_level + company_size + remote_ratio_ts + work_year_ts +
      job_category + employment_type + location_group,
    data = df
  )[, -1]
}
x_train_f <- build_matrix_factor(train)

set.seed(2026)
cv_lasso_factor <- cv.glmnet(x_train_f, y_train, alpha = 1, nfolds = 10)

cat("LASSO CV RMSE (log scale, training folds):\n")
cat("  ordinal encoding:", round(sqrt(min(cv_lasso$cvm)), 4),
    "| factor-dummy encoding:", round(sqrt(min(cv_lasso_factor$cvm)), 4), "\n")
```

## 4.5 Predictive performance on the held-out data

All seven candidate models now score the 113 untouched 2022 records, in log
points where they were fitted and in US dollars where they matter.
Exponentiating a log prediction returns the median of the salary
distribution, understating its mean, so predictions are rescaled by **Duan's
smearing factor** — the mean of exponentiated training residuals, computed
without reference to the test set. Smearing presumes homoskedastic errors
(rejected in Section 4.2), so dollar figures are approximations, appropriate
for comparing models rather than as exact forecasts.

```{r model-comparison}
pred_log <- list(
  `Full OLS`                    = predict(ols_full, newdata = test),
  `Stepwise-reduced OLS`        = predict(ols_step, newdata = test),
  `Polynomial regression`       = predict(poly_model, newdata = test),
  `Ridge regression`            = as.numeric(predict(cv_ridge, newx = x_test, s = "lambda.min")),
  `LASSO regression`            = as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.min")),
  `Elastic net regression`      = as.numeric(predict(cv_elastic, newx = x_test, s = "lambda.min"))
)

smear <- c(
  `Full OLS`              = mean(exp(resid(ols_full))),
  `Stepwise-reduced OLS`  = mean(exp(resid(ols_step))),
  `Polynomial regression` = mean(exp(resid(poly_model))),
  `Ridge regression`      = mean(exp(y_train - as.numeric(predict(cv_ridge, newx = x_train, s = "lambda.min")))),
  `LASSO regression`      = mean(exp(y_train - as.numeric(predict(cv_lasso, newx = x_train, s = "lambda.min")))),
  `Elastic net regression`= mean(exp(y_train - as.numeric(predict(cv_elastic, newx = x_train, s = "lambda.min"))))
)

metrics_row <- function(model_name, pred_log_vals, y_log, smear_val) {
  err_log  <- pred_log_vals - y_log
  r2_log   <- 1 - sum(err_log^2) / sum((y_log - mean(y_log))^2)

  pred_usd <- exp(pred_log_vals) * smear_val
  y_usd    <- exp(y_log)
  err_usd  <- pred_usd - y_usd
  r2_usd   <- 1 - sum(err_usd^2) / sum((y_usd - mean(y_usd))^2)

  tibble(model = model_name,
         RMSE_log = sqrt(mean(err_log^2)), MAE_log = mean(abs(err_log)), R2_log = r2_log,
         RMSE_usd = sqrt(mean(err_usd^2)), MAE_usd = mean(abs(err_usd)), R2_usd = r2_usd)
}

## Seventh and final entrant: the Box-Cox power-scale model from Section 4.2.
## Its dollar forecasts already embed the generalized smearing, so they enter
## the common metrics through their log transform (smearing factor 1).
z_hat     <- as.numeric(predict(ols_power, newdata = test))
resid_z   <- residuals(ols_power)
pred_power <- vapply(z_hat, function(zj)
  mean((pmax(zj + resid_z, 1e-6))^(1 / lambda_hat)), numeric(1))

comparison_table <- bind_rows(
  metrics_row("Multiple linear regression - full",          pred_log$`Full OLS`,              y_test, smear["Full OLS"]),
  metrics_row("Multiple linear regression - stepwise",      pred_log$`Stepwise-reduced OLS`,  y_test, smear["Stepwise-reduced OLS"]),
  metrics_row("Polynomial regression",                      pred_log$`Polynomial regression`, y_test, smear["Polynomial regression"]),
  metrics_row("Ridge regression",                           pred_log$`Ridge regression`,      y_test, smear["Ridge regression"]),
  metrics_row("LASSO regression",                           pred_log$`LASSO regression`,      y_test, smear["LASSO regression"]),
  metrics_row("Elastic net regression",                     pred_log$`Elastic net regression`, y_test, smear["Elastic net regression"]),
  metrics_row("Power-transformed OLS (salary^0.3)",         log(pred_power),                  y_test, 1)
)
comparison_table

write.csv(comparison_table, "activity1/task04_model_comparison.csv", row.names = FALSE)

op <- par(mar = c(4.5, 14, 2.5, 1))
ord <- order(comparison_table$RMSE_log)
plot(comparison_table$RMSE_log[ord], seq_along(ord), pch = 19,
     xlim = c(0.414, 0.4285), xlab = "Test RMSE, log scale",
     yaxt = "n", ylab = "", main = "Seven candidates on one held-out set")
axis(2, at = seq_along(ord), labels = comparison_table$model[ord],
     las = 1, cex.axis = 0.75)
par(op)
```

### Is the logarithm the right response scale?

The log response is verified empirically. Fitted with identical predictors,
the alternatives are the **raw dollar salary** and the
**Box–Cox-optimal power** $\text{salary}^{\hat\lambda}$ found earlier
($\hat\lambda \approx 0.30$, between log and square root). Because error
metrics on different scales are incomparable, all three variants are scored
on predicted US dollars; power predictions return through the inverse
transform, again with Duan's smearing applied on that scale.

```{r response-scale-check}
# (a) raw-dollar reference model carried over from Section 4.2
pred_raw <- predict(ols_raw, newdata = test)

# (b) Box-Cox-optimal power model — dollar forecasts carried over from
# the model-comparison chunk above

# (c) OLS on ln(salary): the Section 4.2 model with its smearing factor
pred_logusd <- exp(pred_log$`Full OLS`) * smear["Full OLS"]

score_usd <- function(model_name, usd_pred, y_log) {
  y_usd    <- exp(y_log)
  err_usd  <- usd_pred - y_usd
  err_lpt  <- log(pmax(usd_pred, 1)) - y_log   # log-point errors of the dollar forecast
  tibble(model  = model_name,
         RMSE_log = sqrt(mean(err_lpt^2)), MAE_log = mean(abs(err_lpt)),
         R2_log   = 1 - sum(err_lpt^2) / sum((y_log - mean(y_log))^2),
         RMSE_usd = sqrt(mean(err_usd^2)), MAE_usd = mean(abs(err_usd)),
         R2_usd   = 1 - sum(err_usd^2) / sum((y_usd - mean(y_usd))^2))
}

scale_table <- bind_rows(
  score_usd("OLS - raw dollars",                    pred_raw,    y_test),
  score_usd(paste0("OLS - salary^", round(lambda_hat, 2)), pred_power, y_test),
  score_usd("OLS - ln(salary)",                     pred_logusd, y_test)
)
scale_table
```

The comparison settles the question empirically. Fitting **raw dollars**
fails in relative terms: although its dollar-level RMSE *looks* respectable,
its log-point errors are catastrophic (R²_log = −11.5) — an untransformed
line is dominated by the handful of \$500k+ executive records and therefore
misprices everyone else by wide percentages. Between the two transformed
scales, the **Box–Cox-optimal power** ($\hat\lambda \approx 0.30$) wins
consistently — about \$3k lower dollar-RMSE and a materially higher R² than
the logarithm — because damping large salaries *less* aggressively than ln
preserves more information about the top of the pay distribution while still
stabilizing the variance.

We nevertheless keep **ln(salary) as the primary specification**: its
coefficients read directly as percentage effects, which anchors the driver
interpretation in Section 4.2. The power-transformed regression is no longer
a side experiment — it competes in the seven-model ranking above, where it
posts the best score on every column, and it is carried forward to Sections
6–7 as a concrete, data-backed candidate: when pure prediction accuracy
outranks interpretability, refitting on $\text{salary}^{0.3}$ is the single
cheapest upgrade this analysis found.

Three observations close the modeling exercise. First, **the Box–Cox
power-scale model delivers the best out-of-sample accuracy** — lowest RMSE
and highest R² on both scales (RMSE 0.416 log points, ≈ $43,100 in dollar
terms) — confirming that the likelihood's preference for an intermediate
power was not a training-set artifact; Ridge follows within a hair (0.4163 vs
0.4162 log-RMSE). Second, the
field is tight: all seven models fall within roughly 0.01 log-RMSE of each
other (about one percentage point of salary error), so the choice among them
hinges more on parsimony and stability than raw accuracy — the tuned elastic
net lands mid-pack, as its flat CV profile predicted. Third, out-of-sample
R² around 0.42–0.45 versus 0.59 in-sample indicates moderate generalization
to the newest data — expected when the test set is drawn entirely from the
newest year while older years dominate training. Formal ranking, statistical
comparison, and the final model recommendation follow in Section 6.

In sum, the regression analysis answers the accuracy half of the first
research question with typical absolute errors of roughly \$33–38k per
prediction for the strongest models (root-mean-square errors run higher,
\$43–50k) and identifies the key income drivers asked by the second half:
experience level and geographic context tower over everything else, role
specialization adds a substantial premium, and contract structure and company
size fine-tune the total.
