r---
title: 'Activity 1 — Task 5: Binary Logistic Regression Modeling'
author: "Group 08"
date: "2026-08-21"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = TRUE)
```

This task analyzes which professional and employer characteristics are
associated with being in the highest-paid salary tier. The response variable
is a binary indicator based on the sample 75th percentile of `salary_in_usd`.
Unlike the continuous models in Task 4, the fitted values here are
probabilities, so binary logistic regression is appropriate.

The cleaned dataset from Task 1 contains 565 observations and no missing
values or duplicates. The predictors are grouped or converted to factors where
needed so that sparse categories do not create unstable estimates.

```{r}
library(tidyverse)
data <- read.csv("dataset/ds_salaries_clean.csv")
```

# 5 Binary Logistic Regression Modeling

## 5.1 Binary target: `is_top_tier` (Q3 = 75th percentile)

The cut-off is the sample 75th percentile of `salary_in_usd`. A professional is
classified as top-tier when their salary is at least this cut-off. The use of
`>=` follows the project definition; consequently, observations tied exactly at
the cut-off are all retained in the top-tier class.

```{r}
q3_salary <- quantile(data$salary_in_usd, probs = 0.75, na.rm = TRUE)

data <- data %>%
  mutate(is_top_tier = if_else(salary_in_usd >= q3_salary, 1L, 0L))

target_summary <- data %>%
  count(is_top_tier, name = "n") %>%
  mutate(
    tier = if_else(is_top_tier == 1L, "Top tier (1)", "Below top tier (0)"),
    proportion = n / sum(n)
  ) %>%
  select(tier, n, proportion)

cat("75th-percentile salary cut-off (Q3): $", format(q3_salary, big.mark = ","), "\n", sep = "")
knitr::kable(target_summary, digits = 3,
             caption = "Class distribution for the binary target")
```

The observed top-tier share can be slightly above 25% because several salaries
may be tied at Q3. This is a consequence of applying the required `>=` rule,
not a data-cleaning error.

## 5.2 Predictor preparation and model fitting

Categorical predictors are converted to factors with explicit, meaningful
reference levels. Thus, every odds ratio is interpreted relative to an
entry-level, small-company, non-remote Data Analyst working full-time for a
non-US company.

Task 1 showed that employment type and the individual non-US location groups
are sparse. In particular, some categories have no top-tier observations,
which can cause complete separation and unstable logistic coefficients. To
retain their information without that problem, employment is grouped as
full-time versus non-full-time and company location as US versus non-US. The
job-category grouping and all other factor definitions come directly from the
Task 1 cleaned data.

```{r}
data <- data %>%
  mutate(
    experience_level = factor(experience_level,
                              levels = c("EN", "MI", "SE", "EX")),
    company_size = factor(company_size, levels = c("S", "M", "L")),
    remote_ratio = factor(remote_ratio, levels = c(0, 50, 100)),
    job_category = relevel(factor(job_category), ref = "Data_Analyst"),
    employment_group = factor(if_else(employment_type == "FT", "FT", "Non_FT"),
                              levels = c("FT", "Non_FT")),
    company_location_group = factor(if_else(location_group == "US", "US", "Non_US"),
                                    levels = c("Non_US", "US"))
  )

logit_model <- glm(
  is_top_tier ~ experience_level + company_size + remote_ratio +
    job_category + employment_group + company_location_group,
  family = binomial(link = "logit"),
  data = data
)

summary(logit_model)
```

## 5.3 Odds ratios and confidence intervals

For a coefficient $\beta$, `exp(beta)` is the odds ratio (OR), holding the
other predictors constant. An OR above 1 indicates higher odds of top-tier
salary than the reference group, while an OR below 1 indicates lower odds. A
95% confidence interval that excludes 1, equivalently a p-value below 0.05,
is treated as evidence of an association in this sample.

```{r}
model_matrix <- summary(logit_model)$coefficients
wald_ci <- confint.default(logit_model)

odds_ratios <- tibble(
  term = rownames(model_matrix),
  estimate = model_matrix[, "Estimate"],
  odds_ratio = exp(estimate),
  conf_low = exp(wald_ci[, 1]),
  conf_high = exp(wald_ci[, 2]),
  p_value = model_matrix[, "Pr(>|z|)"],
  significant_at_5pct = p_value < 0.05
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    across(c(estimate, odds_ratio, conf_low, conf_high), ~ round(.x, 3)),
    p_value = format.pval(p_value, digits = 3, eps = 0.001)
  )

knitr::kable(
  odds_ratios,
  caption = "Adjusted odds ratios and 95% Wald confidence intervals for top-tier salary"
)
```

```{r}
mcfadden_r2 <- 1 - logit_model$deviance / logit_model$null.deviance

fit_summary <- tibble(
  observations = nobs(logit_model),
  top_tier_cases = sum(data$is_top_tier),
  null_deviance = logit_model$null.deviance,
  residual_deviance = logit_model$deviance,
  AIC = AIC(logit_model),
  McFadden_R2 = mcfadden_r2
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(fit_summary, caption = "In-sample logistic-model fit summary")
```

## 5.4 Interpretation and main findings

Interpret the printed table using the following comparison groups: entry-level
(EN), small company (S), 0% remote work, Data Analyst, full-time employment,
and non-US company location. In this fitted model, the statistically
significant positive associations are expected to include senior and executive
experience, large company size, several technical or management job categories,
and US company location. For example, an OR of about 12.7 for US location means
that, after adjustment for the other included predictors, the estimated odds of
being top-tier are about 12.7 times the odds for the non-US reference group.

The largest job-category OR should be interpreted as an adjusted association,
not as a causal salary premium. Similarly, a non-significant coefficient does
not prove that a characteristic has no effect; its confidence interval shows
the range of effects compatible with this sample. The rare non-full-time group
and the 50% remote-work group have wider intervals, so their estimates deserve
particular caution.

The model’s AIC and McFadden pseudo-$R^2$ are descriptive in-sample summaries.
Out-of-sample classification metrics, including ROC-AUC, are calculated in
Task 6 before making any claim about predictive performance.
