---
title: 'Activity 1 — Task 5: Binary Logistic Regression Modeling'
author: "Group 08"
date: "2026-08-21"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE)
```

This task analyzes which professional and employer characteristics are
associated with being in the highest-paid salary tier. The response variable
is a binary indicator based on the sample 75th percentile of `salary_in_usd`.
Unlike the continuous models in Task 4, the fitted values here are
probabilities, so binary logistic regression is appropriate.

The cleaned dataset from Task 1 contains 565 observations and no missing
values or duplicates. The predictors are grouped or converted to factors where
needed so that sparse categories do not create unstable estimates.

```{r}
library(tidyverse)
data <- read.csv("dataset/ds_salaries_clean.csv")
```

# 5 Binary Logistic Regression Modeling

## 5.1 Binary target: `is_top_tier` (Q3 = 75th percentile)

The cut-off is the sample 75th percentile of `salary_in_usd`. A professional is
classified as top-tier when their salary is at least this cut-off. The use of
`>=` follows the project definition; consequently, observations tied exactly at
the cut-off are all retained in the top-tier class.

```{r}
q3_salary <- quantile(data$salary_in_usd, probs = 0.75, na.rm = TRUE)

data <- data %>%
  mutate(
    row_id = row_number(),
    is_top_tier = if_else(salary_in_usd >= q3_salary, 1L, 0L)
  )

target_summary <- data %>%
  count(is_top_tier, name = "n") %>%
  mutate(
    tier = if_else(is_top_tier == 1L, "Top tier (1)", "Below top tier (0)"),
    proportion = n / sum(n)
  ) %>%
  select(tier, n, proportion)

cat("75th-percentile salary cut-off (Q3): $", format(q3_salary, big.mark = ","), "\n", sep = "")
knitr::kable(target_summary, digits = 3,
             caption = "Class distribution for the binary target")
```

The observed top-tier share can be slightly above 25% because several salaries
may be tied at Q3. This is a consequence of applying the required `>=` rule,
not a data-cleaning error.

## 5.2 Training and testing partitions

Task 4 uses a year-aware split, so this task keeps all 2020–2021 records in
training and samples 165 records from 2022 for training. The remaining 113
2022 records are held out for testing.

```{r split-5-2}
set.seed(2026)
train_2022_ids <- sample(which(data$work_year == 2022), 165)
train_ids <- c(which(data$work_year %in% c(2020, 2021)), train_2022_ids)
train <- data[train_ids, ]
test <- data[-train_ids, ]

stopifnot(nrow(train) == 452, nrow(test) == 113, all(test$work_year == 2022))
```

## 5.3 Predictor preparation and model fitting

Categorical predictors are converted to factors with explicit, meaningful
reference levels. Thus, every odds ratio is interpreted relative to an
entry-level, small-company, non-remote Data Analyst working full-time for a
non-US company.

Task 1 showed that employment type and the individual non-US location groups
are sparse. In particular, some categories have no top-tier observations,
which can cause complete separation and unstable logistic coefficients. To
retain their information without that problem, employment is grouped as
full-time versus non-full-time and company location as US versus non-US. The
job-category grouping and all other factor definitions come directly from the
Task 1 cleaned data.

```{r}
job_category_levels <- levels(
  relevel(factor(data$job_category), ref = "Data_Analyst")
)

prepare_predictors <- function(df) {
  dplyr::mutate(
    df,
      experience_level = factor(
        df$experience_level,
        levels = c("EN", "MI", "SE", "EX")
      ),
      company_size = factor(
        df$company_size,
        levels = c("S", "M", "L")
      ),
      remote_ratio = factor(
        df$remote_ratio,
        levels = c(0, 50, 100)
      ),
      job_category = factor(
        df$job_category,
        levels = job_category_levels
      ),
      employment_group = factor(
        dplyr::if_else(df$employment_type == "FT", "FT", "Non_FT"),
        levels = c("FT", "Non_FT")
      ),
      company_location_group = factor(
        dplyr::if_else(df$location_group == "US", "US", "Non_US"),
        levels = c("Non_US", "US")
      )
  )
}

train <- prepare_predictors(train)
test <- prepare_predictors(test)

logit_model <- glm(
  is_top_tier ~ experience_level + company_size + remote_ratio +
    job_category + employment_group + company_location_group,
  family = binomial(link = "logit"),
  data = train
)

summary(logit_model)
```

## 5.4 Odds ratios and confidence intervals

For a coefficient $\beta$, `exp(beta)` is the odds ratio (OR), holding the
other predictors constant. An OR above 1 indicates higher odds of top-tier
salary than the reference group, while an OR below 1 indicates lower odds. A
95% confidence interval that excludes 1, equivalently a p-value below 0.05,
is treated as evidence of an association in this sample.

```{r}
model_matrix <- summary(logit_model)$coefficients
wald_ci <- confint.default(logit_model)

odds_ratios <- tibble(
  term = rownames(model_matrix),
  estimate = model_matrix[, "Estimate"],
  odds_ratio = exp(estimate),
  conf_low = exp(wald_ci[, 1]),
  conf_high = exp(wald_ci[, 2]),
  p_value = model_matrix[, "Pr(>|z|)"],
  significant_at_5pct = p_value < 0.05
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    across(c(estimate, odds_ratio, conf_low, conf_high), ~ round(.x, 3)),
    p_value = format.pval(p_value, digits = 3, eps = 0.001)
  )

knitr::kable(
  odds_ratios,
  caption = "Adjusted odds ratios and 95% Wald confidence intervals for top-tier salary"
)
```

```{r}
mcfadden_r2 <- 1 - logit_model$deviance / logit_model$null.deviance

fit_summary <- tibble(
  observations = nobs(logit_model),
  top_tier_cases = sum(train$is_top_tier),
  null_deviance = logit_model$null.deviance,
  residual_deviance = logit_model$deviance,
  AIC = AIC(logit_model),
  McFadden_R2 = mcfadden_r2
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(fit_summary, caption = "In-sample logistic-model fit summary")
```

To visualize the logistic regression curve for a model with multiple categorical predictors, we plot the predicted probabilities against the model's linear predictors (log-odds). This creates the characteristic "S-curve" of logistic regression.

The actual observations (0 or 1) are jittered slightly on the y-axis to reduce overplotting and make the density of the data points visible.

```{r predicted-probability-curve, fig.width=8, fig.height=5}
# 1. Extract the linear predictors (log-odds) and fitted probabilities
plot_data <- train %>%
  mutate(
    predicted_prob = predict(logit_model, type = "response"),
    log_odds = predict(logit_model, type = "link")
  )

# 2. Plot the logistic curve and the actual observations
ggplot(plot_data, aes(x = log_odds)) +
  
  # Scatter plot of actual observations (Y = 0 or 1)
  # Jitter is added to the y-axis to match the visual style of the reference plot
  geom_jitter(
    aes(y = is_top_tier), 
    width = 0, 
    height = 0.02,
    color = "orange", 
    alpha = 0.3, 
    size = 2
  ) +
  
  # The S-curve representing the predicted probability
  geom_line(
    aes(y = predicted_prob), 
    color = "blue", 
    linewidth = 1.2
  ) +
  
  # Formatting the plot
  theme_minimal() +
  labs(
    title = "Predicted Probability Curve from the Model",
    x = "Linear Predictor (Log-Odds)",
    y = "Probability of Top-Tier Salary"
  ) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25)) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold")
  )
```

The fitted model is applied to the held-out records below; these probabilities
can be used by Task 6 to calculate classification metrics.

```{r test-predictions}
test_predictions <- test %>%
  mutate(
    predicted_probability = predict(logit_model, newdata = test, type = "response"),
    predicted_class = if_else(predicted_probability >= 0.5, 1L, 0L)
  )

head(test_predictions[c("row_id", "is_top_tier", "predicted_probability", "predicted_class")])
```

## 5.5 Interpretation and main finding

The odds-ratio table is estimated from the 452 training observations. Each odds ratio compares one category with its reference category while holding the other predictors constant. The reference profile is entry-level (EN), small-company (S), 0% remote work, Data Analyst, full-time employment, and a non-US company location. For example, an odds ratio of about 26.4 for US company location means that the estimated odds of being top-tier are 26.4 times those for a comparable non-US company, not that salary itself multiplies by this factor.

Interpret the printed table using these comparison groups. In this fitted model, the statistically significant positive associations include senior and executive experience, large company size, several technical or management job categories, and US company location. Notably, the odds ratio for executive-level experience is extraordinarily high (over 253), although its wide confidence interval suggests some instability due to a small sample size in this specific sub-group. Other strong predictors include Senior level (OR $\approx$ 30.8) and ML/AI Engineer roles (OR $\approx$ 20.6) compared to entry-level Data Analysts.

These are adjusted associations, not causal salary effects. The model contains no interaction terms, so each coefficient describes an average association under the model’s additive assumption. A non-significant coefficient does not prove that a characteristic has no effect; it simply indicates a lack of statistically significant evidence at the 5% level in this sample. Estimates for the non-full-time and 50% remote-work groups should be treated cautiously because their smaller sample sizes produce wider confidence intervals.

  The AIC and McFadden pseudo-$R^2$ summarize in-sample fit on the training data. The `test_predictions` object contains predictions for the 113 held-out records; classification metrics calculated from those predictions, including ROC-AUC, should be evaluated in Task 6 before making any claims about out-of-sample predictive performance.