---
title: "Activity 1 — Task 7: Interpretation and Recommendations"
author: "Group 08"
date: "2026-08-30"
output:
  html_document:
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This task closes the modeling loop of Activity 1. Task 6 already selected a
"best" model for each research question; here we do two things the requirement
asks for. First, we interpret which features actually drive Data Science pay
(§7.1) and check whether the three independent methods used — continuous
regression, binary logistic regression, and factorial ANOVA — agree on the same
drivers. Second, we propose concrete avenues for improving predictive
performance (§7.2), each grounded in a limitation that our own model outputs
exposed rather than in generic machine-learning advice.

We deliberately reuse the exact partitions, encodings, and results of Tasks 3–6.
No model is re-evaluated on the held-out test set here beyond what Tasks 4–6
already did; the recommender narrative below simply reads the numbers that our
own fitted models produced.

```{r load}
library(tidyverse)
library(broom)
library(knitr)
library(pROC)

data  <- read.csv("dataset/ds_salaries_clean.csv")
train <- read.csv("dataset/train_data.csv")
test  <- read.csv("dataset/test_data.csv")
cmp   <- read.csv("task04_model_comparison.csv")
```

# 7 Interpretation and Recommendations

## 7.1 Feature importance interpretation

### 7.1.1 One model to predict, one model to interpret

Task 6 chose the **Ridge regression** as the best continuous predictor on the
held-out test set (the Box–Cox power model finished a hair behind), and the
**binary logistic model** as the best classifier for the top-25% target. The
Ridge winner achieves its accuracy by shrinking coefficients and therefore has
no clean coefficient table to read. For feature-importance interpretation we
deliberately read its in-sample sibling: the **log-salary regression** selected
by AIC stepwise in Task 4 (§4.2). Its coefficients translate directly into
percentage effects via $100(e^{\hat\beta}-1)$ — a view that neither the Ridge
winner nor the power model offers on their respective penalized and
$\text{salary}^{0.26}$ scales. Ridge and the stepwise OLS share the same
predictors and the same training records, so reading the stepwise table for
substance does not change the story.

The context table from Task 6 is repeated here so every number cited below is
traceable:

```{r metrics-recap}
# Recompute the Task 6 out-of-sample metrics on the same partition so this
# recap cannot drift from Task 6's own printed output.
q3_salary <- as.numeric(quantile(data$salary_in_usd, probs = 0.75, na.rm = TRUE))

prepare_metrics_data <- function(df) {
  df %>%
    mutate(
      is_top_tier = if_else(salary_in_usd >= q3_salary, 1L, 0L),
      experience_level = factor(experience_level, levels = c("EN", "MI", "SE", "EX")),
      company_size = factor(company_size, levels = c("S", "M", "L")),
      remote_ratio = factor(remote_ratio, levels = c(0, 50, 100)),
      job_category = relevel(factor(job_category), ref = "Data_Analyst"),
      employment_group = factor(if_else(employment_type == "FT", "FT", "Non_FT"),
                                levels = c("FT", "Non_FT")),
      company_location_group = factor(if_else(location_group == "US", "US", "Non_US"),
                                      levels = c("Non_US", "US"))
    )
}
train_for_metrics <- prepare_metrics_data(train)
test_for_metrics  <- prepare_metrics_data(test)

logit_metrics <- glm(is_top_tier ~ experience_level + company_size + remote_ratio +
                       job_category + employment_group + company_location_group,
                     family = binomial(link = "logit"), data = train_for_metrics)

test_probs <- predict(logit_metrics, newdata = test_for_metrics, type = "response")
test_preds <- if_else(test_probs >= 0.5, 1L, 0L)
actual     <- test_for_metrics$is_top_tier

tn_m <- sum(test_preds == 0L & actual == 0L); fp_m <- sum(test_preds == 1L & actual == 0L)
fn_m <- sum(test_preds == 0L & actual == 1L); tp_m <- sum(test_preds == 1L & actual == 1L)
acc_m <- (tp_m + tn_m) / length(actual)
prec_m <- tp_m / (tp_m + fp_m)
rec_m  <- tp_m / (tp_m + fn_m)
f1_m   <- 2 * prec_m * rec_m / (prec_m + rec_m)
auc_m  <- as.numeric(auc(roc(actual, test_probs, quiet = TRUE)))

logistic_metrics <- tibble(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "F1-Score", "ROC-AUC"),
  Value  = round(c(acc_m, prec_m, rec_m, f1_m, auc_m), 3)
)

knitr::kable(cmp %>% arrange(RMSE_usd),
             caption = "Continuous candidates ranked by out-of-sample USD RMSE (Task 6)")
knitr::kable(logistic_metrics,
             caption = "Out-of-sample classification metrics of the logistic model (Task 6)")
```

### 7.1.2 Recovering the driver tables

To interpret the selected models we refit them *on the same training partition
that Tasks 4 and 6 used* (452 rows, seed 2026) and reconstruct the two tables
that the interpretation builds on: the percentage effects of the stepwise
log-salary model and the odds ratios of the logistic model.

```{r refit-continuous, echo=TRUE}
# Reproduce the Task 4 factor reference levels
train$experience_level <- factor(train$experience_level, levels = c("EN", "MI", "SE", "EX"))
train$company_size     <- factor(train$company_size,     levels = c("S", "M", "L"))
train$employment_type  <- relevel(factor(train$employment_type), ref = "FT")
train$job_category     <- relevel(factor(train$job_category),   ref = "Data_Analyst")
train$location_group   <- relevel(factor(train$location_group), ref = "Other")

ols_full <- lm(log_salary ~ experience_level + job_category + employment_type +
                 company_size + location_group + remote_ratio + work_year + is_local,
               data = train)
ols_step <- step(ols_full, direction = "both", trace = FALSE)   # Task 4's $4.2 model

pct_effects <- broom::tidy(ols_step) %>%
  filter(term != "(Intercept)") %>%
  mutate(effect_pct = 100 * (exp(estimate) - 1),
         p_val      = format.pval(p.value, digits = 3, eps = 0.001)) %>%
  select(term, estimate, p_val, effect_pct)
knitr::kable(pct_effects %>% arrange(desc(effect_pct)),
             digits = 3,
             caption = "Stepwise log-salary model: coefficients and percentage effects (reference: EN / Data Analyst / FT / Small / Other-location)")
```

```{r refit-logistic, echo=TRUE}
q3_salary <- quantile(data$salary_in_usd, probs = 0.75, na.rm = TRUE)   # $150,000

prepare_logit <- function(df) {
  df %>%
    mutate(
      is_top_tier = if_else(salary_in_usd >= q3_salary, 1L, 0L),
      experience_level = factor(experience_level, levels = c("EN", "MI", "SE", "EX")),
      company_size = factor(company_size, levels = c("S", "M", "L")),
      remote_ratio = factor(remote_ratio, levels = c(0, 50, 100)),
      job_category = relevel(factor(job_category), ref = "Data_Analyst"),
      employment_group = factor(if_else(employment_type == "FT", "FT", "Non_FT"),
                                levels = c("FT", "Non_FT")),
      company_location_group = factor(if_else(location_group == "US", "US", "Non_US"),
                                      levels = c("Non_US", "US"))
    )
}
train_li <- prepare_logit(train)

logit <- glm(is_top_tier ~ experience_level + company_size + remote_ratio +
               job_category + employment_group + company_location_group,
             family = binomial(link = "logit"), data = train_li)

or_table <- tibble(
  term      = rownames(summary(logit)$coefficients),
  estimate  = summary(logit)$coefficients[, "Estimate"],
  odds_ratio = exp(estimate),
  p_val     = summary(logit)$coefficients[, "Pr(>|z|)"]
) %>%
  filter(term != "(Intercept)") %>%
  mutate(odds_ratio = round(odds_ratio, 2),
         p_val      = format.pval(p_val, digits = 3, eps = 0.001))
knitr::kable(or_table,
             caption = "Logistic model: coefficients and odds ratios for top-25% salary (reference: EN / Data Analyst / FT / Small / non-US)")

cat("Class balance, training:", table(train_li$is_top_tier),
    "| McFadden R2:", round(1 - logit$deviance / logit$null.deviance, 3), "\n")
```

### 7.1.3 Cross-method agreement on the key drivers

The three independent methods — log-scale percentage effects, logistic odds
ratios, and the heteroskedasticity-robust ANOVA from Task 3 — are collected in
the table below. Asterisks mark estimates significant at the 5% level in their
own model.

```{r agreement-table}
agreement <- tribble(
  ~Driver,                                  ~"Regression (% vs ref)", ~`Logistic OR`,   ~`Robust ANOVA (Task 3)`,
  "Experience: Mid (MI)",                   "+24%*",                    "n.s.",          "main effect",
  "Experience: Senior (SE)",                "+65%*",                    "18.0*",         "p < 0.001*",
  "Experience: Executive (EX)",             "+157%*",                   "137.5*",        "(all levels)",
  "Company location: US",                   "+155%*",                   "19.1*",         "not in ANOVA",
  "Residence match (is_local)",             "+51%*",                    "not in model",  "not in ANOVA",
  "Job: Data Engineer",                     "+33%*",                    "6.8*",          "not in ANOVA",
  "Job: Data Scientist",                    "+39%*",                    "6.1*",          "not in ANOVA",
  "Job: ML / AI Engineer",                  "+50%*",                    "20.2*",         "not in ANOVA",
  "Job: Management / Lead",                 "+59%*",                    "11.0*",         "not in ANOVA",
  "Job: Research",                          "+70%*",                    "13.6*",         "not in ANOVA",
  "Company size: Medium (M)",               "+17%*",                    "3.3*",          "n.s. (robust p = 0.19)",
  "Company size: Large (L)",                "+35%*",                    "6.7*",          "n.s. (robust p = 0.19)",
  "Employment: Part-time",                  "-67%*",                    "grouped (n.s.)","not in ANOVA",
  "Employment: Freelance",                  "-66%*",                    "grouped (n.s.)","not in ANOVA",
  "Remote ratio",                           "retained (n.s.)",          "n.s.",          "not in ANOVA",
  "Work year",                              "dropped (n.s.)",           "n.s.",          "2022 surge* (unadjusted)"
)
knitr::kable(agreement,
             caption = "Cross-method agreement on salary drivers. * = significant at 5% within that model; n.s. = not significant; grouping of sparse employment categories follows Task 5; is_local (residence matches employer country) is part of the Task 4 regression but not of the logistic or ANOVA designs.")
```

Three conclusions stand out from the table.

**Experience level is the dominant driver of pay.** In the regression, moving
from entry-level to mid pays roughly +24%, to senior +65%, and to executive
+157% — a steep, monotone gradient. The logistic model agrees: senior and
executive levels raise the odds of top-25% pay by 18- and 137-fold relative to
entry level. The robust ANOVA agrees as well, confirming experience as the
strongest factor ($p < 0.001$) *after* heteroskedasticity is corrected, while
every other tested effect weakens or disappears. The executive estimates carry
very wide confidence intervals (the OR of 137.5 has a 95% interval from about
17 to 1,084) because few executive observations exist; they are best read as
"very large" rather than taken literally.

**Geography is the second major driver, and it is structural.** A US-based
employer is worth roughly +155% in salary — the single largest coefficient in
the model — with Canada (+88%), Germany (+68%), Great Britain (+69%) and France
(+37%) forming a clear second tier and India about −52% below the rest-of-world
baseline. The logistic model captures the same fact in coarse form: US location
multiplies the odds of a top-25% salary by about 19. A related new signal
emerged once `is_local` entered the regression: being employed where you live is
worth roughly **+51%** over an otherwise identical posting at an overseas
employer — consistent with the cross-country gap being structural rather than an
artifact of which countries happen to appear in the sample. The size of these
effects is what a country-level dummy can capture; they almost certainly bundle
together several things (local wage levels, cost of living, tax regimes,
seniority mixes) that a finer data set could split apart.

**Role family matters, within the field.** Against the Data-Analyst baseline,
engineering (+33%) and data science (+39%) pay moderately more, while
machine-learning/AI (+50%), management (+59%) and research (+70%) pay
substantially more. The logistic odds ratios replicate the same ordering
(6.1–20.2). The heterogeneous "Other" bucket is significant in the logistic
model but not in the regression — the two methods simply disagree about a
catch-all category with no economic meaning, which is itself a reason to refine
the grouping rather than trust either verdict.

**The remaining factors are secondary and the methods diverge on company
size.** Regression and logistic both find significant, positive company-size
contrasts (medium +17%, large +35%; OR 3.3 and 6.7), yet the robust ANOVA finds
company size *not* significant once unequal variances are corrected
(p = 0.19) and the medium-size contrast is only borderline under the
regression's own robust errors (p = 0.083). Part-time (−67%) and freelance
(−66%) salaries are clearly penalized, but these rest on 10 and 4 records
respectively, and in the logistic model the folded "non-full-time" group is not
significant at all. Remote work ratio and survey year, significant in isolation
(the 2022 surge from Task 3), carry no stable incremental signal once the other
factors are held constant — stepwise dropped the survey year (keeping
`remote_ratio`, itself insignificant), and LASSO shrank the year component to
exactly zero.

### 7.1.4 Where the methods disagree

The one genuine disagreement is Company Size, and it is *structural* rather
than economic. Task 3 showed that the unadjusted two-way ANOVA finds a
significant experience-by-size interaction (p = 0.008) which evaporates under
the heteroskedasticity-robust correction (p = 0.13), while the size main effect
also loses significance (p = 0.19). Yet the regression and logistic models,
fitted on the same data with robust training of their own, still report
significant size coefficients. The reason: regression and logistic are also
affected by the non-constant variance that the ANOVA correction removes. The
prudent read — and the one we adopt — is that experience, geography and role
family are **robust** drivers, while company size is a *suggestive* driver whose
apparent influence partly mirrors the unequal spread of salaries across groups.
Any claim about company size should therefore be presented as conditional, not
definitive.

## 7.2 Avenues for model improvement

### 7.2.0 Why a plateau, not a failure

The results of Task 6 carry an honest limitation: the seven continuous
candidates sit within about 0.009 log-RMSE of one another (0.580–0.588) and
within about \$2.9k of USD RMSE. In other words, *switching model families buys
almost nothing*. Meanwhile, out-of-sample $R^2$ on the log scale (about
0.34–0.36) trails the in-sample adjusted $R^2$ of the log model (0.64) by a wide
margin, and residual diagnostics in Task 4 (§4.2) rejected normality on every
response scale tested — with heteroskedasticity persisting on both the log
(BP = 54.7, p = 3·10⁻⁴) and power (BP = 42.2, p = 0.012) scales.

Together these three facts say the remaining error is not a property of the
linear model family but of **how the problem is represented**: the data's
granularity, its measurement quality, and how we validate on a single 113-row
draw. The improvement avenues below therefore emphasize better data, better
features, and honest validation over "a harder model."

### 7.2.1 Data-side improvements

| Ave. | Avenue | Evidence hook from our analysis | Expected benefit |
|---|---|---|---|
| a | Add more years (2023+) and more records | Only 3 years sampled (n = 565); the 2022 surge in Task 3 (§3.4) means the newest year is a single snapshot; our test set is one draw | Captures the post-2022 hiring-boom correction instead of extrapolating a single year; shrinks metric uncertainty |
| b | Enrich predictors: education, years of experience, industry, employer tier, city-level geography | US is +155% / OR 19.1 — a country-level coarse average that bundles many economic forces | Splits the single largest coefficient into interpretable pieces |
| c | Fix sparse groups at the source rather than by grouping | 36 of 53 employer countries have ≤ 3 rows (Task 1); part-time (10), contract (5), freelance (4); Task 5 had to fold non-FT together | Fewer information-destroying groupings; keeps granular levels in the model |
| d | Improve target quality | `salary` is self-reported gross pay; the USD conversion uses yearly-mean FX rates (fxdata.foorilla.com); no hours worked, so PT/FL understate effort-adjusted pay | Cleaner coupling of "salary" to what a worker actually earns |

### 7.2.2 Feature-engineering improvements

| Ave. | Avenue | Evidence hook from our analysis | Expected benefit |
|---|---|---|---|
| e | Treat `remote_ratio` as its three categories instead of one linear 0–100 slider | It takes only values {0, 50, 100}; stepwise keeps it as a linear term while the quadratic adds nothing (F = 2.48, p = 0.085, Task 4 §4.3) | A three-category dummy can still recover a distinct premium at the 50% hybrid point that a single linear term cannot represent |
| f | Test interaction terms deliberately | The only ANOVA-tested interaction (Exp × Size) was fragile — p = 0.008 unadjusted, p = 0.13 robust (Task 3 §3.3); Exp × Role and Year × Experience are untested | Cannot yet claim "interactions do not matter"; a targeted test would settle it |
| g | Refine `is_local` into residence-by-remoteness pieces (e.g., `is_local` interacted with `remote_ratio`) | `is_local` now enters the regression and is worth +51% (Task 4 §4.2), but as one blunt 0/1 flag it cannot separate "works abroad" from "employer abroad" | Splits the residence premium into a remote-work component and a tax/effort component, replacing one average with two interpretable ones |
| h | Revisit the response scale per model — ln for reading, salary^0.26 for prediction, or a Gamma GLM with log link | Box–Cox interval [0.22, 0.34] excludes both λ = 1 and λ = 0 (Task 4 §4.2); heteroskedasticity persists on both the log and the power scales | A principled alternative to "pick a power and hope"; Gamma suits multiplicative, positive-skewed pay |

### 7.2.3 Modeling and validation improvements

| Ave. | Avenue | Evidence hook from our analysis | Expected benefit |
|---|---|---|---|
| i | Mixed-effects model with employer country as a random effect | Task 1 lumped 46 sparse countries into `Other`; small countries plausibly share a latent level | Borrows strength across countries instead of discarding level information |
| j | Quantile regression on log-salary (e.g., 75th percentile) | The whole classification problem is the top 25%; OLS/ANOVA model conditional *means* and are pulled by heavy tails (Task 3 §3.3) | Models exactly the pay range RQs 2 and 3 care about; robust to heteroskedastic tails |
| k | Tree ensembles / gradient boosting with heavy cross-validation | Residual non-normality persists on every scale (Task 4 §4.2); trees capture interactions linear models flatten | Non-linear capture plus a second source of feature-importance (SHAP) to back up §7.1 |
| l | Robust estimation, not just robust standard errors (WLS or Huber/M-type) | HC-robust errors fix inference but the coefficients themselves are still OLS (Task 4 §4.2) | Estimates that down-weight the influential, high-variance cells ANOVA flagged |
| m | Ensemble / stack the seven candidates | All seven sit within ~0.009 log-RMSE — each holds different information | Cheap expected gain by averaging models that agree on the plateau |
| n | Uncertainty-aware validation: walk-forward and repeated k-fold, bootstrap CIs on metrics | Our "winner" (Ridge) beats the power transform by ~$0.2k in USD RMSE — inseparable from noise on a single 113-row holdout (Task 6 §6.1) | Replaces point estimates with intervals; tells us which gaps are real |

### 7.2.4 Prioritization and the honest boundary

We rank the fourteen avenues by expected benefit against effort, and by which
limitation each fixes:

```{r priorities}
priorities <- tribble(
  ~Priority, ~Avenue,                                  ~Fixes,                      ~Effort,
  "Highest payoff",  "Richer predictors (b) + more recent data (a)", "The dominant US/+155% driver is currently one coarse block; the test set is one year", "High (data collection)",
  "Highest payoff",  "Uncertainty-aware validation (n)",            "Winner-vs-runner-up gaps near $0.2k are noise on one 113-row draw", "Low",
  "Cheapest wins",   "Response-scale work and Gamma GLM (h)",       "Persistent heteroskedasticity and non-normality on both transformed scales", "Low",
  "Cheapest wins",   "Ensemble of the existing seven (m)",          "The model plateau: every candidate within 0.009 log-RMSE", "Low",
  "Moderate value",  "Quantile regression (j) / mixed effects (i) / tree ensembles (k)", "Heavy tails, coarse country grouping, flattened interactions", "Medium",
  "Only if adopted for Activity 2", "Feature engineering (e, f, g) and robust estimation (l)", "Hybrid remote, untested interactions, OLS coefficient estimates", "Medium-High"
)
knitr::kable(priorities,
             caption = "Prioritized improvement road-map (a–n refer to the tables above)")
```

We deliberately recommend *not* rushing to a more exotic model family first: the
seven existing candidates already disagree by only pennies on the dollar, so the
next model will inherit the same ceiling. The two moves with the largest
expected impact are entirely data-side — collecting more recent, richer records
so the giant country effect can be decomposed — and validation-side, so that the
tiny gaps we currently report as rankings are expressed as the uncertainty
intervals they actually are.

Closing honestly: with a self-reported, single-source Kaggle sample of 565
rows, we have strong cross-method evidence that **the drivers are robust**
(experience, geography, role family), but the error *around* those drivers is
idiosyncratic — no modeling treatment we tried removes it. That asymmetry is
exactly what the report's conclusions will carry into Activity 2, where we aim
to test whether a richer data source lets the same methods do better.